Parkinson’s disease-associated Pink1 loss disrupts vesicle trafficking in Ensheathing glia causing dopaminergic neuron synapse loss

Version 1

Abstract

Abstract

Parkinson’s disease (PD) is commonly associated with the loss of dopaminergic neurons in the substantia nigra, but many other cell types are affected even before neuron loss occurs. Recent studies have linked oligodendrocytes to early stages of PD, though their precise role is still unclear. Pink1 is mutated in familial PD and through unbiased single-cell sequencing of the entire brain of Drosophila Pink1 models, we observed significant gene deregulation in ensheathing glia (EG); cells that share functional similarities with oligodendrocytes. We found that the loss of Pink1 leads to the activation of EG, similar to the reactive response of EG seen upon nerve injury. Using cell-type specific transcriptomics, we identified deregulated genes in EG as potential functional modifiers. Specifically, downregulating two trafficking factors, Rab7 and Vps13, also mutated in PD, or the direct regulators of Rab7, Mon1 and Ccz1, specifically in EG was sufficient to rescue neuronal function and protect against dopaminergic synapse loss. Our findings demonstrate that Pink1 loss in neurons triggers an injury response in EG, and that Pink1 loss in EG in turn disrupts neuronal function. Vesicle trafficking components, which regulate membrane interactions between organelles within EG, play a crucial role in maintaining neuronal health and preventing dopaminergic synapse loss. Our work highlights the essential role of glial support cells in the pathogenesis of PD and identifies vesicle trafficking within these cells as a key point of convergence in disease progression.

Introduction

Parkinson’s disease (PD) is characterized by motor symptoms such as bradykinesia, rigidity and tremor that are caused by the progressive loss of dopaminergic neurons in the substantia nigra (Lang & Lozano, 1998). However, most of the patients report non-motor symptoms such as constipation, hyposmia and sleep defects even before the onset of motor symptoms (Munhoz et al., 2015). This suggests that PD is a progressive neurodegenerative disease that initiates many years before the diagnosis and involves different neuronal systems, multiple anatomical areas and different cell types. While therapies symptomatically improve motor functioning temporarily by restoring dopaminergic tone, they do not effectively prevent neurodegeneration and disease progression. This highlights the necessity to identify cell-types and biological mechanisms that act at the earliest stages of disease.

With the advent of single-cell sequencing applied to PD and control brains in combination with evidence from genome-wide association studies (GWAS), it is now possible to suggest which cells and mechanisms are at play at early stages of PD. Among these, a recent study reported that genes related to GWAS loci are highly enriched for oligodendrocyte-specific gene expression (Bryois et al., 2020). Interestingly, differential gene expression in post-mortem brains with Braak scores of 1-2, 3-4, and 5-6 and controls, suggested a role for oligodendrocytes at an early stage of PD prior to the overt loss of dopaminergic neurons(Bryois et al., 2020). Additionally, two independent single-cell sequencing studies, one of the human substantia nigra and one of the midbrain, both associated PD genetic risk with oligodendrocyte-specific expression patterns (Agarwal et al., 2020; Smajic et al., 2022). While these studies started to link dysfunction of oligodendrocyte-cells to the early stages of PD, they did not address their role in the pathophysiology of the disease.

Here we describe a new Drosophila model to assess how this neuron-glia cross-talk contributes to PD. Drosophila has been successfully used to investigate cellular and molecular dysfunction preceding the onset of age-related symptoms, including dopaminergic neuron-dependent motor symptoms (Kaempf et al., 2024; Pech et al., 2024; Valadas et al., 2018). Drosophila glial cells also display several anatomical and functional features that show remarkable similarity to their mammalian counterparts (Freeman & Doherty, 2006; Kremer et al., 2017). While the small size of the Drosophila brain does not necessitate elaborate myelination, ensheathing glia (EG) in this species share not only anatomical features with oligodendrocytes, such as the ability to wrap around nerve tracts in the CNS (Kremer et al., 2017; Yildirim et al., 2019), but also important molecular and functional features, such as providing metabolic support and regulating neuronal activity (Delgado et al., 2018; Otto et al., 2018).

In a brain-wide single-cell sequencing experiment, we identified EG as the most deregulated cell type in young (pre-motor) Pink1 loss-of-function Drosophila model. Using cell-specific labeling, immunohistochemistry and electrophysiological recordings, we correlated the transcriptional deregulation in EG with an “active cell state” also seen upon nerve injury, and showed that healthy EG are fundamental to support dopaminergic synapse integrity in Pink1 mutants. Finally, using cell-type-specific transcriptomic and a high-throughput screening and immunohistochemistry, we identified Rab7 and Vps13 as EG-expressed modifiers that maintain dopaminergic synapse-integrity. Our study shows that neuron-ensheating glia crosstalk is disrupted already at young age by the loss of Pink1 in neurons but also in EG and that EG are fundamental to support dopaminergic synapse-integrity, suggesting a role for oligodendrocytes in the progression of PD.

Results

Ensheating glia in Pink1 loss of function flies are activated non-cell autonomously

We previously created a whole-brain single-cell RNAseq dataset from different Drosophila PD knock-in models. This dataset was generated from young 5-day-old flies to capture ‘early’ changes (Pech et al, 2024). We re-analyzed the data from Pink1P399L knock-in loss-of-function mutant flies and isogenetic controls, and performed differential gene expression (DEG) analysis for each cell type using DESeq2 (Figure 1A). Glial cell subtypes show the highest degree of deregulation; particularly ensheathing glia are strongly affected (Figure 1A). While flies do not myelinate neurons, ensheating glia (EG) serve similar supporting functions to oligodendrocytes (Otto et al., 2018; Yildirim et al., 2019). These results suggest that ensheating glia are deregulated in young Pink1P399Lmutants (and also Pink1KO, see below) at an age prior to dopaminergic neuron-dependent motor defects (Kaempf et al., 2024; Pech et al., 2024).

(A) tSNE of the cells of Pink1P399L knock-in mutants (5 days old). Cell types are labeled with colors indicating the number of deregulated genes compared to control. EG are encircled and labeled. (B-B”) (B) SUM intensity of confocal images of fly brains (5 ± 1 days old) stained with anti-GFP (Green) and anti-Brp (Magenta), where anti-GFP marks EG and anti-Brp marks presynaptic sites of the antennal lobes. Scale bar: 20 µm (B’) SUM intensity of confocal images of control flies vs control flies 24 hours after ORNs severing. (B”) SUM intensity of confocal images of Pink1KO-WS vs Pink1KO-WS flies 24 hours after ORNs severing. (C) Quantification of GFP intensity with the glomeruli of the antennal lobe area in 5 ± 1 days old flies (as in B) relative to the control. ANOVA with Dunnet’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥13 per genotype. (D-D”) SUM intensity of confocal images of fly brains (5 ± 1 days old) stained with anti-GFP (Green) and anti-Brp (Magenta), where anti-GFP marks EG and anti-Brp marks presynaptic sites of the antennal lobes. Scale bar: 20 µm. (D’) SUM intensity of confocal images of w1118 with Pink1 downregulation in EG. (D”) SUM intensity of confocal images of Pink1KO-WSwith Pink1 rescued in EG. (E) Quantification of GFP intensity within the glomeruli of the antennal lobe area in 5 days old flies (as in D) relative to the control. ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥10 per genotype.

When neurons are damaged, activated EG invade the neuropil (Doherty et al., 2009; MacDonald et al., 2006). For example, when fly antennal lobes are removed, severing the olfactory Olfactory Receptor Neurons (ORNs), EG invade the antennal lobe neuropil (MacDonald et al., 2006), potentially as a protective response (Figure 1B’-C). We tested if EG invade the neuropil in Pink1 mutant flies and expressed a transmembrane fluorescent protein (UAS-CD8GFP) under the control of a promoter specific to EG (MZ709-Gal4) in control and Pink1KOflies. Samples were labeled with anti-GFP and the pre-synaptic protein Bruchpilot (NC82). Similar to ORN-severed controls, EG invade the antennal neuropil in Pink1KO, and severing the ORN in Pink1KO does not further exacerbate this phenotype (Figure 1B”-C). These data suggest that Pink1 loss triggers EG activation, similar to neuron-severing, to invade the ORN neuropil.

In order to determine whether this activation is cell-autonomous, we generated cell-type specific Pink1 perturbations. We either downregulated Pink1 using a previously well-validated RNAi line specifically in EG (EG-specific Pink1 loss-of-function – Figure 1D’), or we re-expressed wild-type Pink1 specifically in EG in Pink1KOflies (all cells, but the EG, loss-of-function - Figure 1D”). When Pink1 is knocked-down in EG (but present in other cells, including neurons), EG do not invade the neuropil (Figure 1D’-1E). However, when Pink1 is not present in neurons and expressed in EG, the EG do invade the antennal neuropil (Figure 1D”-E). This suggests that Pink1-defective neurons activate the EG in a cell-non autonomous fashion to invade, and potentially protect, the neuropil.

Ensheathing glia function supports synaptic integrity

Our finding that Pink1KO in neurons elicits an EG activation response, suggested that EG might in turn also functionally modulate neuronal integrity in mutant flies. To test this we made use of a readily accessible model circuit in the fly visual system with an easy electrophysiological read-out, Electroretinograms (ERG). The ERG waveforms are susceptible to changes in intracellular signaling, neuronal function and synaptic transmission and have been well characterized, and amply used to assess neuronal and cellular function(Hardie & Raghu, 2001; Praschberger et al., 2023; Soukup et al., 2016; C. F. Wu & Wong, 1977). We exploited this method to understand the role of Pink1 in EG on neuronal function and synaptic transmission in young flies.

Comparing control with young Pink1KO flies does not show a difference in the “depolarization-response”, indicating that -as expected-there is no neurodegeneration occurring in the photoreceptors at this stage (Figure 2A). However, when comparing the ON peak – which represents synaptic transmission from the photoreceptors and within the underlying neuronal circuit – we detect a significant reduction in the mutant flies compared to the control, suggesting that in Pink1KO flies synaptic communication in this circuit is partially impaired, already at 5 days old (Figure 2A-B).

(A) Representative ERG traces: ON peak is highlighted by the arrow. (B) Normalized ON peak response of flies (5 ± 1 days old). ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥11 per genotype. (C-C”) (C) Maximum projection confocal image of control and Pink1KO-WS in Mushroom Bodies (MB) of aged flies (22 ± 2 days old), stained with anti-TH (cyan) and anti-DLG (magenta) antibodies, DLG is used to mark post-synaptic site of MB. The black and white image is the middle Z-plan within the region of interest of the MB(ROI, yellow), it is used to represent the thresholded TH area (white). Scale bar: 20 µm. (C’) Maximum projection confocal image of w1118 with Pink1 downregulation in EG. (C”) Maximum projection confocal image of Pink1KO-WS with Pink1 rescued in EG. (D) Quantification of dopaminergic synaptic area within MB in aged flies (22 ± 2 days old). ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥12 per genotype.

In order to discern the role of EG in this defective ERG response, we (1) downregulated Pink1 specifically in EG and (2) we expressed Pink1 specifically in EG of Pink1KO. When Pink1 is knocked down specifically in EG, we observe the same reduction in ON peak response as in the Pink1KO flies. This indicates that EG integrity is necessary for efficient neurotransmission in this circuit. Conversely, the ON peak defect of Pink1KO mutants is rescued when Pink1 is reintroduced in EG only. This suggests that activated EG are able to at least partially buffer the detrimental effects of Pink1 deficiency in neurons (Figure 2A-B).

To also test EG function in relation to PD-relevant dopaminergic (DA) synapses we evaluated the loss of dopaminergic neuron afferents. We previously showed that the loss of Pink1 function causes a progressive loss of protocerebral anterior medial dopaminergic neuron (PAM DAN) afferents in the Mushroom Body neuropil (Kaempf et al., 2024) and confirm here that 5 days old Pink1KO flies do not show a decrease in the dopaminergic synaptic area in the mushroom bodies (MB) lobes, while aged, 25 days old animals, do (Figure 2 C-D). Hence, while ensheathing glia are already activated at young age in Pink1KO, DA synapses are structurally still intact and only deteriorate in older flies. We then knocked down Pink1 specifically in EG and find a similar strong decrease in DA synaptic area (Figure 2C’-D). Conversely, when Pink1 is reintroduced specifically in EG of Pink1KO animals, DA synapse-loss is significantly rescued (Figure 2C”-D). Together, these results indicate that Pink1 in EG is necessary to protect dopaminergic neurons from synapse loss.

EG cell type-specific transcriptomics reveals modifiers of neuronal dysfunction

To identify pathways in EG that are deregulated by Pink1KO loss-of-function, we optimized a method to isolate EGs specifically from the rest of the brain, enabling us to perform much higher-sensitivity transcriptomics than what we achieved with our 10x droplet-based single-cell sequencing (Pech et al., 2024). We expressed a fluorescently-tagged histone (UAS-HisTag-eGFP) in EG (GMR-56-gal4) and used Fluorescence-activated Cell Sorting (FACS) to isolate EG. We sorted 300 cells per technical replicate, and performed Bulk-RNA sequencing using a SMART-seq2 protocol (Figure 3A). The sequencing results of EG-sorted cells and of neuron-sorted cells (nsyb-Gal4 instead of GMR-56-gal4) show strong enrichment of EG cell markers in the former and strong enrichment of neuronal markers in the latter (Figure 3B). We then compared the transcriptomic profile of EG in Pink1KO to control using differential expression analysis and find 617 deregulated genes in EG (Appendix 1, padj<0.05). To understand whether the deregulated genes were enriched for a specific pathway or associated with a cellular component compared to non-significantly deregulated genes, we performed GO analysis of the deregulated genes in Pink1KOflies. Our analysis resulted in generic terms and very diverse pathways that did not allow us to draw further conclusions.

(A) Scheme of cell-type specific transcriptomics. Panel created using BioRender.com/p56u250. (B) Scaled gene expression of representative genes for EG and neurons after sorting EG or neurons using the protocol described in figure A. Figure generated using R studio (R Core Team, 2022). (C) Differentially expressed genes in EG in Pink1KO-WScompared to control flies, plotted according to their Log2 Fold change and the -Log10 of the adjusted p-value. Intercept in RED (-Log10 adjusted p-value = 4.31). *Two data points are outside the boundaries of the plot. (D) ERG ON peak value differences to Pink1KO-WS (red) flies (5 ± 1 days old) of Pink1KO-WS flies (5 ± 1 days old) with DEGs downregulated, or upregulated, specifically in EG. ON peak value are expressed as the difference to Pink1KO-WS. ANOVA with Dunnett’s test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥3 per genotype. *One data point is outside the boundaries of the plot.

To test if the genes deregulated in EG are modifiers of the neuronal pink1KO phenotypes, we downregulated and upregulated (when possible) the 50 most deregulated genes, for which we could find RNAi fly lines, specifically in the EG of Pink1KO and recorded ERGs in 5 days old flies. Several of the genes show a rescue of the ON transient defects in Pink1KO (4 were significant suppressors at p<0.05) and few exacerbate the defect (Figure 3D). Among the 4 significant suppressors of the Pink1KOphenotypes, 3 (Ccz1, CG10646, CG17660) are associated with endosomal membranes - or their human ortholog is - (Figure 3C-D) (Borchers et al., 2023; Epp et al., 2011; Gaudet et al., 2011; Gershlick et al., 2016). Hence, several DEG in EG of Pink1KO are modifiers of the Pink1 ERG mutant phenotype, and most of those are associated with the endosome.

Rab7 and Vps13 downregulation in EG rescue synaptic dopaminergic neurons afferent loss

Among the genes associated with the endosome that we identified in our ERG screening, the strongest suppressors, Ccz1, is associated with late endosome function. In particular, Ccz1 forms together with Mon1 a guanosine-nucleotide-exchange factor (GEF) that recruits Rab7 to the endosome (Borchers et al., 2023; Epp et al., 2011). Interestingly, Rab7 is known to physically interact with a peripheral membrane protein that is also causative of PD, Vps13 (Vonk et al., 2017). We therefore also tested if Rab7, Mon1, and Vps13 are also Pink1-modifiers and find that the knockdown of each of these genes also rescues the ON transient defect of Pink1KO mutants (Figure 4A’-B).

(A-A’) (A) Representative ERG traces. (A’) Representative ERG traces of genes Pink1KO-WS flies with Vps13, mon1, and rab7 downregulated in EG (B) Normalized ON peak response of flies (5 ± 1 days old). Kruskal-Wallis with Dunn’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥8 per genotype. (C) Maximum projection confocal image of control and Pink1KO-WS in Mushroom Bodies (MB) of aged flies (22 ± 2 days old), stained with anti-TH (cyan) and anti-DLG (magenta) antibodies, DLG is used to mark post-synaptic site of MB. The black and white image is the middle Z-plan within the region of interest of the MB(ROI, yellow), it is used to represent the thresholded TH area (white). Scale bar: 20 µm. (D) Quantification of dopaminergic synaptic area within MB in aged flies (22 ± 2 days old). ANOVA with Dunnett’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Bars: mean ± SD; points are individual animals N≥13 per genotype.

To determine if manipulation of vesicle trafficking specifically in EG can rescue PAM DAN synaptic innervation in old (20-25 day old) Pink1KO mutants, we expressed RNAi to Ccz1, Vps13, Rab7 or Mon1 using mz709-Gal4. Anti-TH labeling (marking DAN synapses) in the Mushroom Body area (marked by anti-DLG) is significantly reduced in Pink1KO, but rescued to control levels when any of the four genes are knocked down (Figure 4D). These results indicate that Rab7-mediated membrane trafficking in EG regulate glia-neuron crosstalk to maintain synaptic integrity of dopaminergic neuron synapses in the fly brain of Pink1 mutants.

Discussion

In this work, we show there is early, non-cell-autonomous activation of EG in a PD-relevant Drosophila Pink1 model. This occurs already in young Pink1-mutant flies when neuronal defects in dopaminergic neurons are not yet measurable. The EG activation and invasion phenotype appears as a protective response, as Pink1-deficient neurons seem to drive EG activation in a manner that mimics the response to nerve injury (Doherty et al., 2009; MacDonald et al., 2006). In addition to EG activation, we show there is also a non-autonomous role of EG to support Pink1-deficient neurons. When we manipulate the expression of membrane-lipid trafficking genes, including the homolog of the established PD gene VPS13C (Lesage et al., 2016), specifically in EG, we rescue the dysfunction of Pink1 mutant neurons. Our work shows the importance of neuron-glia cross-talk in the context of Pink1-deficiency. We find (1) there is early EG activation secondary to Pink1-loss-induced neuronal impairments; (2) there are specific lipid- and membrane trafficking problems caused by Pink1 loss in EG that center on Rab7-signaling and (3) there is a convergence of Parkinson-relevant genetic factors that act in EG to maintain neuronal function.

The loss of Pink1 function in the context of Parkinson’s disease has been amply linked to the regulation of mitochondrial health. Pink1 maintains the integrity of the electron transport chain by phosphorylating NDUFA10 (Morais et al., 2009, 2014; Pogson et al., 2014) and when mitochondria are damaged, it phosphorylates Parkin and Ubiquitin to facilitate mitophagy (Kane et al., 2014; Narendra et al., 2010; Narendra & Youle, 2024). The regulation of mitophagy is complex and requires mitochondrial rearrangements controlled by mitochondria-organelle contacts (ER and lysosomes) (Wong et al., 2018, 2019; Yamano et al., 2018). Such contacts mediate inter-organellar lipid exchange as well as facilitate organellar fusion (Kumar et al., 2018; Valadas et al., 2018). In Pink1 mutants these contacts are increased in number and this causes cellular defects (Grossmann et al., 2023; Valadas et al., 2018).

We found that in EG, Rab7 and genes encoding proteins that regulate the Rab7 GTPase, functionally interact with Pink1-induced neuronal dysfunction: their genetic manipulation in EG rescues cell non-autonomous neuronal dysfunction. Loss of Rab7 itself, or loss of Mon1 and Ccz1, only in EG are sufficient to rescue neuronal Pink1 phenotypes. Mon1 and Ccz1 form a guanine exchange factor (GEF) that specifically loads Rab7 with GTP, thereby activating the protein (Borchers et al., 2023). This is interesting because Rab7-GTP promotes the formation of mitochondria-organelle contacts (mitochondria-lysosome contacts) (Wong et al., 2018). This is also further supported by previous observations showing that overexpression of Fis1 rescues Pink1 mutant phenotypes: Fis1 localizes to mitochondria to recruit TBC1D15 (Wong et al., 2018), a Rab7 GTPase Activating Protein that hydrolyzes GTP bound to Rab7, thereby inactivating the protein and promoting untethering of lysosome-mitochondrial contacts (Wong et al., 2018). Hence, conditions that inhibit Rab7-GTP-mediated mitochondria-organelle contacts (eg by inhibiting Rab7-GEF or activating Rab7-GAP) in EG rescue Pink1 neuronal dysfunction, possibly by negatively regulating the stability of organelle contacts in Pink1 mutants.

Our observation that downregulation of Vps13 in EG also rescues Pink1 is in further support of a role for organelle membrane contact site regulation. Vps13 physically interacts with Rab7 (Vonk et al., 2017) and it has two mammalian homologs, VPS13A and VPS13C (Vonk et al., 2017; Vrijsen et al., 2022), that is also mutated in familial PD (Lesage et al., 2016). VPS13A tethers ER to mitochondria and lipid droplets and VPS13C tethers ER to late endosomes, lysosomes and lipid droplets (Kumar et al., 2018; Muñoz-Braceras et al., 2019; Vrijsen et al., 2022). The function of these proteins is to facilitate lipid transfer between organelles, a function we previously showed to be increased in Pink1 mutants (Valadas et al., 2018). Hence, the downregulation of Vps13, similar to Rab7 or its regulators, would counteract the effects of excessive organelle membrane contact formation and dysregulated lipid homeostasis in Pink1 mutants.

Lipid homeostasis in glial cells is crucial for supporting neuronal function by providing energy, protecting against oxidative stress, and regulating synaptic health. In oligodendrocytes, maintaining myelin sheaths is critical (Ettle et al., 2016; Lappe-Siefke et al., 2003). While fly EG do not generate such structures, they do wrap processes around neuropil regions, also requiring lipid membrane production (Kremer et al., 2017; Pogodalla et al., 2021). Through lipid metabolism, glia also supply neurons with essential metabolites, cholesterol, and signaling molecules that aid in membrane integrity, synaptic remodeling, and neuroprotection (Delgado et al., 2018; Otto et al., 2018; Saab et al., 2016; Suárez-Pozos et al., 2020). Further work is now required to define which functions supported by organelle contact sites in glial cells drive cell-nonautonomous protective mechanisms in neurons. Our work in flies indicates these glial defects precede dopaminergic synapse loss, and, importantly, that rescuing the glial defects rescues the dopaminergic problems later in life, underscoring the significance for therapeutic development.

Methods

Resource availability

Lead Contact

Further information and requests for resources and reagents should be directed to the lead contact, Patrik Verstreken (patrik.verstreken@kuleuven.be).

Materials Availability

Data, code, Drosophila models, and reagents are available upon request.

Fly stocks

The fruit flies were maintained in an incubator at 25°C under a 12:12 light-dark cycle, and provided with a standard diet consisting of cornmeal and molasses. For experiments, only fly males were used. Flies were raised in parallel on the same batch of food, which was exchanged every three to four days and kept in only male population of similar density. Flies were aged to 5 ± 1 days old, and 22 ± 2 days old as indicated for immunohistochemistry and ERGs respectively.

To create a set of Drosophila models for Parkinson’s disease, CRISPR/Cas9-based gene editing was utilized, as outlined in (Kaempf et al., 2024; Pech et al., 2024). In brief, each knockout line contained an attP-flanked w+ reporter cassette that replaced the first shared exon across different isoforms of the targeted gene. For this study, we used knockout flies for Pink1 from this collection (Kaempf et al., 2024). The “control” strain denotes a semi-isogenized w1118 strain that underwent backcrossing to Canton-S for 10 successive generations, resulting in a strain termed Canton-S-w1118. To generate the fly line Yw; ; UAS-His2Av::eGFP.VK27/TM3Sb, we created the pUASTattB_His2AV-GFP plasmid, linearizing the pUAST.attB (Bischof et al., 2007) with EcoRI-BamHI. A gBlock, containing the His2AV, GS linker and eGFP sequence, was cloned into this linearized plasmid with Gibson Assembly. The plasmid was inserted into the VK27 landing site by BestGene. The sequence His2AV-GS-eGFP gblock is:TGAATAGGGAATTGGGcAAacATGGCTGGCGGTAAAGCAGGCAAGGATTCGGGCAAGGCCAAGGCGAAGG CGGTATCGCGTTCCGCGCGCGCGGGTCTTCAGTTCCCCGTGGGTCGCATCCATCGTCATCTCAAGAGCCGCACT ACGTCACATGGACGCGTCGGAGCCACTGCAGCCGTGTACTCCGCTGCCATATTGGAATACCTGACCGCCGAGG TCCTGGAGTTGGCAGGCAACGCATCGAAGGACTTGAAAGTGAAACGTATCACTCCTCGCCACTTACAGCTCGC CATTCGCGGAGACGAGGAGCTGGACAGCCTGATCAAGGCAACCATCGCTGGTGGCGGTGTCATTCCGCACAT ACACAAGTCGCTGATCGGCAAAAAGGAGGAAACGGTGCAGGAcCCGCAGCGGAAGGGCAACGTCATTCTGTC GCAGGCCTACGGTTCAGGCGGAGGTGGCAGCGGCGGTGGCGGATCCATGGTGAGCAAGGGCGAGGAGCTG TTCACCGGGGTGGTGCCCATCCTGGTCGAGCTGGACGGCGACGTAAACGGCCACAAGTTCAGCGTGTCCGGC GAGGGCGAGGGCGATGCCACCTACGGCAAGCTGACCCTGAAGTTCATCTGCACCACCGGCAAGCTGCCCGTG CCCTGGCCCACCCTCGTGACCACCCTGACCTACGGCGTGCAGTGCTTCAGCCGCTACCCCGACCACATGAAGCA GCACGACTTCTTCAAGTCCGCCATGCCCGAAGGCTACGTCCAGGAGCGCACCATCTTCTTCAAGGACGACGGC AACTACAAGACCCGCGCCGAGGTGAAGTTCGAGGGCGACACCCTGGTGAACCGCATCGAGCTGAAGGGCATC GACTTCAAGGAGGACGGCAACATCCTGGGGCACAAGCTGGAGTACAACTACAACAGCCACAACGTCTATATCA TGGCCGACAAGCAGAAGAACGGCATCAAGGTGAACTTCAAGATCCGCCACAACATCGAGGACGGCAGCGTGC AGCTCGCCGACCACTACCAGCAGAACACCCCCATCGGCGACGGCCCCGTGCTGCTGCCCGACAACCACTACCT GAGCACCCAGTCCGCCCTGAGCAAAGACCCCAACGAGAAGCGCGATCACATGGTCCTGCTGGAGTTCGTGACC GCCGCCGGGATCACTCTCGGCATGGACGAGCTGTACAAATAAGGGTACCTCTAGAGGATCTT.

Flies were backcrossed for five generations into an inbred Canton-S strain harboring the w1118 mutations. The following genotypes were used in this study:

Olfactory Receptor Neurons (ORNs) axotomy

ORN bilateral axotomy was performed on 4 ± 1 days old flies by surgical ablation of the third antennal segments. Briefly, flies were anesthetized with CO2 and then both antennae were removed with Dumont #5 forceps, and returned to the vial (M. Purice, 2020; M. D. Purice et al., 2017). Flies recovered 24 hours on food at 25°C before dissection.

Immunohistochemistry and confocal imaging – invasion phenotype

Immunohistochemistry was performed on adult flies brains of 5 ± 1 days, with at least two independent experiments. The flies The brains were dissected in ice-cold PBS and fixed for 20 minutes in freshly prepared 3.7% paraformaldehyde (in 1x PBS, 0.3% Triton X-100 [PBX] [Sigma Aldrich]) at room temperature (RT), followed by three 15-minute washes in PBX at RT on a shaker. Then, the brains were incubated for 1 hour in blocking solution (PBX, 10% normal goat serum [NGS]) at RT. Following blocking, the brains were incubated with primary antibodies (rabbit anti-GFP [Thermo Fisher Scientific], 1:1000, mouse anti-brp [DSHB, nc82], 1:100) in blocking solution at 4°C overnight, followed by three 15-minute washes in PBX at RT. Secondary antibodies (goat anti-rabbit Alexa488, goat anti-mouse Alexa555, both at 1:1000 [Thermo Fisher Scientific]) in PBX with 10% NGS were applied overnight at 4°C. Afterward, the brains were washed three times for 15 minutes in PBT at RT on a shaker and mounted with the anterior facing up in RapiClear 1.47 (SUNJin Lab). Imaging was performed using a Nikon A1R confocal microscope with a 40x (NA 1.15) water immersion lens, Z-stacks of the entire antennal lobes were acquired. The acquisition was carried out using a Galvano scanner, with a zoom factor of 1, scan speed of 0.5, and line averaging set to 2. All images were captured with a pinhole of 0.9 Airy units and a resolution of 1024 × 1024. Z-stacks (with 0,5 μm step intervals) were obtained for data acquisition, and the same imaging settings were applied across all genotypes and sessions. Image analysis was performed using Fiji (Schindelin et al., 2009). Image analysis was performed using Fiji (Schindelin et al., 2009). Each antennal lobe is composed of multiple glomeruli, and EG separate them and invade them when active (Doherty et al., 2009; Pogodalla et al., 2021). To quantify EG invasion in the antennal lobe, we calculated the invasion in each single glomeruli. To do so and be consistent with the section of the antennal lobe analyzed, anti-brp was used to identify the 3 Z-planes where DM6 and DM1 glomeruli were present (B. Wu et al., 2017). Then to automatically detect and segment each glomeruli of the antennal lobe section, regions of interest (ROIs) were defined in the SUM projection of the 3 Z-planes by using the Fiji plugin Stardist, which is developed to segment round objects (Weigert et al., 2019). To quantify the EG invasion in the ROIs, anti-GFP SUM projection of the same 3 Z-planes was used. The GFP intensity was detected in each ROI, and it was normalized to the area of each ROI (GFP intensity/glomeruli area) to calculate the EG invasion in each glomeruli. Then, the normalized GFP intensity of each glomeruli of an antennal lobe were summed and normalized for the number of glomeruli detected in the antennal lobe. For each experiment, the EG antennal lobe invasion of each fly was normalized to the mean of the control. Representative images show the SUM projections of the 3 Z-planes.

Immunohistochemistry and confocal imaging – TH staining

Immunohistochemistry was performed on adult fly brains of 22 ± 2 days, with at least two independent experiments. The brains were dissected in ice-cold PBS and fixed for 20 minutes in freshly prepared 3.7% paraformaldehyde (in 1x PBS, 0.2% Triton X-100 [PBX]) at room temperature (RT), followed by three 15-minute washes in PBX at RT on a shaker. Then, the brains were incubated for 1 hour in blocking solution (PBX, 10% normal goat serum [NGS]) at RT. Following blocking, the brains were incubated with primary antibodies (rabbit anti-TH [Sigma], 1:200, mouse anti-DLG [DSHB], 1:100) in blocking solution at 4°C for 1.5 to 2 days, followed by three 15-minute washes in PBX at RT. Secondary antibodies (goat anti-rabbit Alexa488, goat anti-mouse Alexa555, both at 1:500 [Thermo Fisher Scientific]) in PBX with 10% NGS were applied overnight at 4°C. Afterward, the brains were washed three times for 15 minutes in PBT at RT on a shaker and mounted with the anterior facing up in RapiClear 1.47 (SUNJin Lab). Imaging was performed using a Nikon A1R confocal microscope with a 20x (NA0.95) water immersion lens, Z-stacks of the entire brain were acquired. The acquisition was carried out using a Galvano scanner, with a zoom factor of 1, scan speed of 0.5, and line averaging set to 2. All images were captured with a pinhole of 2.3 Airy units and a resolution of 1024 × 1024. Z-stacks (with 3 μm step intervals) were obtained for data acquisition, and the same imaging settings were applied across all genotypes and sessions. Image analysis was performed using Fiji (Schindelin et al., 2009). To quantify dopaminergic neuron innervation of the mushroom body (MB), anti-DLG was used to identify the five z-planes containing the synaptic region of the MB lobes. The region of interest (ROI) for the MB was defined in the sum projection of the five z-planes. To exclude background signal comparable to control, the area of anti-TH fluorescence was thresholded (using the default threshold for all z-planes) within the selected z-stacks. Quantification of the thresholded area within the ROI was performed in each z-plane, then summed and normalized to the MB ROI area for each brain individually (TH+ area/MB area). For each experiment, the individual TH+ area/MB area values were normalized to the mean of the control. Representative images show the maximum projection of five z-planes and the thresholded middle z-plane.

Electroretinograms (EGR)

ERGs were recorded from flies 5 ± 1 day old as previously described (Heisenberg, 1971; Slabbaert et al., 2016). Flies were immobilized on glass microscope slides using double-sided tape. For recordings, glass electrodes (borosilicate, 1.5 mm outer diameter) filled with 3 M NaCl were placed in the thorax as a reference and on the fly eye for recordings. Each fly was exposed to 5 cycles of 3 seconds of darkness, followed by a 1-second of light stimuli with LED illumination. Response to the stimuli was recorded using Axoscope 10.7 and analyzed using Clampfit 10.7 software (Molecular Devices). ERG traces were analyzed with Igor Pro 6.37 (Wave Metrics) using a custom-made macro.

Single-cell dissociation for cell-type specific transcriptomics

To test that EG could be isolated from the rest of the brain we used two cohorts GMR56F03> UAS-His2Av::eGFP.VK27 and nSyb-Gal4> UAS-His2Av::eGFP.VK27. To identify differentially expressed genes in EG two cohorts were used the control w1118 expressing GMR56F03> UAS-His2Av::eGFP.VK27 and Pink1KO-WS expressing GMR56F03> UAS-His2Av::eGFP.VK27. For each genotype, ten brains of flies 5 ± 1 days old were collected, with most of the laminae removed, from each experimental repeat. The dissections were performed in ice-cold PBS that contained 5 μM Actinomycin D (Sigma) to inhibit changes in gene expression during tissue processing(Y. E. Wu et al., 2017). To prevent any bias caused by different experimental batches, all the genotypes were processed in parallel. To minimize variability across batches, the same reagents and procedures were used throughout the experiment. Dissections were completed within one hour. The dissection order and the assignment of genotypes to dissectors were alternated to reduce any experimenter-related variation.

For the dissociation of brain tissue, a mix of Dispase I (0.6 mg/ml) (Sigma), Collagenase I (30 mg/ml) (Thermo Fisher), Trypsin (0.5x) (Thermo Fisher), and 5 μM Actinomycin D was used. The brains were incubated for 20 minutes at 25°C with 1000 rpm shaking, performing four triturations at 5-minute intervals. After dissociation, the cell suspension was washed with PBS containing 5 μM Actinomycin D and then filtered through a 10 μm strainer (Pluriselect), using 300 μl of PBS EDTA (Sigma) 1μM. DAPI (Invitrogen) (1:300.000) was added after filtration.

Fluorescence Activated Cell Sorting (FACS) for cell-type specific transcriptomics

Immediately after the dissociation, cells were sorted using a FACSAria Fusion (BD Biosciences) flow cytometer equipped with 4 lasers (405nm, 488nm, 561nm and 640nm) and a 100 µm nozzle at 20 psi. Live (DAPI-negative), GFP-positive cells were sorted into 96-well PCR plates at 300 cells per well (3 wells per genotype) using a 4-way purity mask with the FACSDiva software v9.0.1 (BD Biosciences).

Bulk transcriptomics of EG cells

Library preparations and sequencing for the Bulk RNA seq of the EG cells was performed using a modified Smart-seq2 protocol based on a previously published protocol (Picelli et al., 2013). Briefly, as per the protocol, 3 ul of cell lysis buffer (0.1% Triton X-100, Sigma Aldrich; 1U/ul of RNase Inhibitor, RNase OUT, Thermo Fisher; 2.5 uM Oligo dT(25), IDT; and 2.5 mM dNTP, Promega) was aliquoted in triplicate to 96 well plates (4titude, Cat. No. 4TI-0960/C). Roughly 300 cells were sorted into the wells with lysis buffer and the plate was spun down at 2000xg for 1 min prior to storage at - 80*C. For the first strand synthesis, the Smart-Seq 2 plates were thawed to room temperature for a minute and subsequently spun down at 2000 xg for minute. The RNA denaturation is carried out at 72*C for 10 mins and then flash cooled on ice for 5 mins. First strand synthesis mix (1X SuperScript II first strand synthesis buffer, Thermo Fisher; 5 mM DTT, Thermo Fisher; 100 U SuperScript II reverse transcriptase enzyme, Thermo Fisher; 10 U RNAse OUT, Thermo Fisher; 1M Betaine, Sigma Aldrich; 6 mM MgCl2, Thermo Fisher; and 1 uM TSO, IDT Technologies) is added to the cell lysis mix to a total volume of 10 ul. The First strand synthesis reaction was carried out with following program: 42*C for 90 mins; 10 cycles of {50*C for 2 mins; 42*C for 2 mins}; 72*C for 15 mins; 4*C indefinite hold. PCR amplification of the first strand product was performed by adding 15 ul of the PCR mix to the first strand product (1X KAPA HiFi HotStart Ready Mix, Roche; and 0.1 uM of IS PCR primer, IDT Technologies). 22 cycles of the following PCR program was used to amplify the first strand product: 98*C for 3 mins; 22 cycles of {98*C for 20 secs; 67*C for 15 secs; 72*C for 6 mins}; 72*C for 5 mins; 4*C indefinite hold. 20 ul of Ampure XP was added to each well and mixed. Standard Ampure XP purification was carried out as per manufacturer’s recommendation and the cDNA library was eluted in 17.5 ul of Elution buffer. 17 ul of the eluted library was transfered to fresh 96 well plate (4titude, Cat. No. 4TI-0960/C). 5 ul of Tn5 tagmentation mix (1X Tagment DNA buffer, Illumina; ATM mix, Illumina and cDNA library 1 ng) was prepared for cDNA fragmentation. Tn5 tagmentation was performed with following program: 55*C for 10 mins; 4*C indefinite hold. Tagementation reaction was stopped by quenching the reaction with 1.25ul of NT buffer for 5 mins (Illumina). 1.25 ul of the i5 and i7 Illumina indexes were added to the plates. Finally 3.75 ul of NPM master mix was added to the plate and mixed well and put for the index PCR amplification: 72*C for 3 mins; 95*C for 30 secs; 12 cycles of {95*C for 10 secs; 55*C for 30 secs; 72*C for 30 secs}; 72*C for 5 mins, and 4*C indefinite hold.

The indexed PCR products were pooled in a single tube and 0.8X Ampure XP purification (Beckman Coulter) was carried out as per manufacturer’s recommendation and finally the sequencing library was eluted in 35.5 ul of Elution buffer (Qiagen).

SMART seq 2 libraries were sequenced on the NextSeq 500 (Illumina) sequencing platform. Sequencing was done as per the protocol recommendations: paired end read of 76 bps (read 1), 76 bps (read 2), 8 bps (index 1) and 8 bps (index 2). For a targeted sequencing depth of ∼30 million reads per sample.

Analysis of RNAseq data - Neuron vs EG

Raw reads from neuron- vs. glia-specific FACsorting experiments were processed with the nf-core/rnaseq pipeline (Patel et al., 2020) with default parameters and the BDGP6 reference genome. The STAR-aligned and salmon-quantified expression matrix was then subset to genes known to be enriched in neurons or glia and displayed as a per gene scaled heatmap.

Analysis of RNAseq data - EG specific differentially expressed genes

After FACsorting of PinkKO-ws vs. control ensheathing glia, library preparation and sequencing resulting FASTQ files were cleaned with fastp (Chen et al., 2018) with default settings. Aligning and counting were carried out with STAR (Dobin et al., 2013)and the 4th 2020 FlyBase Drosophila melanogaster release (r6.35) with the –quantMode flag set as GeneCounts. For differential gene expression testing we used DESeq2 (Love et al., 2014), where we fit a negative binomial model and carried out the Wald-test. We removed experimental batch as a covariate according to the design formula ∼date+genotype. All genes below a Benjamini-Hochberg corrected p-value of 0.05 were considered deregulated.

ScRNAseq data loading, filtering, clustering, and differential expression analysis

Raw counts data from Pech et al., 2024, alongside cell annotations provided by the authors was loaded into Scanpy (v1.9.1) (Wolf et al., 2018) and subset to only include cells in the young control (<= 6d of age) or Pink1 samples. After subsetting the data, filtering was performed to remove any cell with less than 250 genes expressed or a percentage of counts coming from mitochondrial genes greater than 15%, data was then normalized to a total of 10,000 counts and log transformed. Highly variable genes were detected, total_counts and percent of mitochondrial reads was regressed out and finally the data was scaled to unit variance with a zero mean, clipped to a max of 10.

After data pre-processing, a PCA was performed, Harmony (Korsunsky et al., 2019) was used to correct batches using “sample_id” as a batch key, and 122 components were used for dimensionality reduction and clustering analyses. Leiden clustering was performed with a resolution of 8.0 and using the annotations from Pech et al., 2024, all cells within each cluster were labeled by taking the most common, original annotation per cluster.

Per cluster, a differential analysis was performed. First, pseudobulk counts were generated by summing the counts per cell from all cells per sample, these counts were then used in DESeq2 (v1.44.0) (Love et al., 2014) comparing Control vs Pink1 samples. The total number of genes significantly up or downregulated per cluster (padj <= 0.05 and an abs (log2foldchange) >= 1.5) was counted and plotted.

Statistical analysis

GraphPad Prism was used for visualization and to determine statistical significance. Datasets were tested for normal distribution using the D’Agostino-Person Omnibus and Shapiro-Wilk normality test. For normally distributed dataset ordinary one-way ANOVA was used, followed to correct for multiple comparisons by a post hoc Tukey’s test when comparing all the datasets with each other or Dunnett’s test when comparing all the datasets to a general control. For non-normally distributed datasets Kruskal-Wallis test, followed to correct for multiple comparisons by a post hoc Dunn’s test. Significance levels are defined as * is p<0.05, ** is p<0.01, *** is p<0.001 and ns, not significant. ‘N’ in the legends is used to indicate how many animals were analyzed. Data are plotted as mean ± SD. Specifics on the statistical test used for each analysis are reported in the figure legends.

eLife Assessment

The preliminary data presented in this manuscript seem valuable. The data are currently incomplete and there are numerous technical concerns, some of which may arise from insufficient description of methodologies used.

Reviewer #1 (Public review):

Summary:

This study investigates the impact of Pink1 loss on glial function and neuronal health in a Drosophila model, highlighting the role of mitochondria-organelle contacts and key genes such as Ccz1, Vps13, Mon1, and Rab7. The work provides insights into cellular processes underlying neurodegenerative diseases, with a focus on glia-neuron interactions. While the findings are promising, the study lacks critical controls, detailed mechanistic evidence, and explanatory figures to strengthen its claims.

Strengths:

(1) The study addresses an important topic in neuroscience, exploring the mechanisms of Pink1 loss, which has implications for Parkinson's disease and neurodegeneration.

(2) The focus on mitochondria-organelle contacts and their regulation by Rab7-mediated pathways is novel and provides a potential mechanism for neuronal dysfunction.

(3) The identification of key genes (Ccz1, Vps13, Mon1, Rab7) and their potential roles in Pink1-related pathways adds valuable knowledge to the field.

(4) The manuscript uses a combination of genetic tools, Drosophila models, and functional assays to approach the problem from multiple angles.

Weaknesses:

(1) Specificity of Mz-Gal4: The study lacks validation of Mz-Gal4 specificity, as it may also drive expression in a few neurons or other types of glia. Additional control experiments using nls-GFP with Elav, Repo, or Draper antibody staining or alternative glial drivers would be helpful.

(2) DLG staining is central to the story but is not well-supported by high-resolution Z-stack imaging, which should be included in the supplementary figures.

(3) The manuscript does not confirm whether the candidate RNAi (Ccz1, Vps13, Mon1, Rab7) directly influence Rab7-mediated membrane trafficking or mitochondria-lysosome contacts in Pink1 mutants.

(4) Using ERG as a readout for EG effects in the antenna is not a direct or appropriate assay. Alternative functional assays relevant to antenna glia should be considered.

(5) A graphical explanation of the interactions and functions of the candidate genes in Pink1 KO mutants is missing. This would greatly enhance the manuscript's clarity.

(6) The study lacks details on sample sizes, effect sizes, and reproducibility, which are necessary for robust conclusions.

(7) There are repeated words on page 3 ("olfactory Olfactory Receptor Neurons") and a lack of explanation in Figure 3C regarding the most up-regulated and down-regulated genes and the significance of large red dots.

Reviewer #2 (Public review):

Summary: This study proposes a novel role for ensheathing glia (EG) in a Pink1-model of Parkinson's disease and shows that this cell population exibits the highest number of DEG in a pre-symptomatic stage. In the olfactory system, there seems to be morphological changes in this cell-type that resembles an 'activated' state and the authors further show that the neuronal loss of Pink1 is responsible for this defect. The authors go on to show that manipulation of Pink1 in EG also leads to some defects in the visual system and in the dopaminergic neurons (DAN) that innervate the mushroom body (MB), and performed a screen based on the 'on-transient' defect of the ERG to identify potential genes that may modulate the function of EG in synaptic regulation. They focus on several genes related to Rab7/Vps13, and performed some additional experiments in the visual system and MB to propose the role of vesicle/lipid trafficking in EG as a important factor for PD pathogenesis.

Strengths: The study proposes functional and mechanistic connections between several genes that have been linked to PD (PINK1, VPS13A/C). I feel that the data presented in Figure 1 and Fig3A-C are performed with rigor and are convincing/novel. The selection of Drosophila to study the questions is also a strength and the lab has extensive experiences in this field and model organism.

Weaknesses: There is one fundamental concern I have with the genetic experiments performed in this paper (especially in Fig 3D and Fig4, see major issue #1), and I feel that there is a bit of a disconnect between the EG 'activation' phenotype the author show in the olfactory system and the other two neuronal systems (visual system, MB DAN) that the authors investigate see major issue #2). Also, there are quite a bit of information that is not provided in the manuscript (see major issues #3 and #4), which makes me difficult to judge the rigor and interpretation of several experiments.

Major Concern #1: A number of lines used in this study are referred to as "RNAi" lines but when I look at the actual genotypes of reagents listed in the table in the METHODS section, many are actually NOT RNAi lines. Quite a few lines, including lines that the authors use as RNAi against Ccz1, Rab7 and Mon1, are gRNA lines for the TKO (TRiP-CRISPR knockout) system. While these reagents can theoretically knock-out these genes in somatic cells if used in combination with UAS-Cas9, there is no mention that UAS-Cas9 was used in this work throughout the manuscript. Hence, when these lines are just crossed to GAL4 with or without the Pink1 mutant, they shouldn't be having any effects. Similarly, the strongest hit from their screen was a TOE (TRiP-CRISPR Over Expression) gRNA against PIG-A, which could allow overexpression of PIG-A if there is a UAS-dCas9::VP64. However, I also do not see any mention that such activator was introduced into the crossing scheme. Considering that 3 of the 4 'hits' from their screen are not RNAi lines, I am quite skeptical of the study. Similarly, except for Vps13, all reagents used in Fig4 are TKO gRNA lines. Therefore, if this experiment was conducted without an UAS-Cas9, most of the data shown here are problematic. Also, note that several of the 'RNAi' lines listed in the Table in the METHODS section are actually MiMIC alleles. While some MiMIC lines could function as strong LOF alleles (if they are inserted in the exon or in an intron of the gene in the same orientation as the gene), some of the lines are not expected to affect gene function (e.g. FASN2 and CG17712, MiMICs are in introns and face the opposite orientation). Hence, the rationale of including these reagents in the screen doesn't make much sense. The description of the modifier screen should be much more detailed in the RESULTS and METHODS section and if the UAS-Cas9/dCas9::VP64 transgenes were not introduced when the TKO/TOE reagents were utilized, what can be concluded?

In addition, for the 4 genes that the authors further study in Fig4, there are many other reagents that the authors can use, including mutant alleles, previously characterized RNAi lines (e.g. Vps13) and dominant negative/constitute active lines (e.g. especially for Rab7). The authors should validate their results with independent reagents to really convincingly show that the same conclusions can be drawn for the Vps13/Rab7 related genes since this is the key takeaway message of this paper.

Also, they do not show whether the manipulation of these genes in a wild-type background (they only show what happens in Pink1 mutants) affect ERG and MB DAN synapse morphology. If these manipulations alone dramatically affect these phenotypes, it would be very difficult to interpret their data.

Major Concern #2: In Figure 1, the authors show some morphological evidence that EG are 'activated' in Pink1 mutants, but whether the same phenomenon occurs in the visual system and in the MB is not shown. Since all of the studies in Fig3D and Fig4 are done in the visual system and MB, it is not clear whether the visual system and MB phenotypes are related to 'activation' of EG.

Also, in the RNA-seq data in Fig1A and Fig3C, is there any molecular evidence that EG are indeed 'activated'? The only evidence that the authors show to state that EG are 'activated' in young Pink1 null animals is based on increased CD8::GFP staining in the olfactory system.

The authors cannot draw a strong conclusion that indeed EG are 'activated' based on these data (e.g. perhaps the expression level of CD8::GFP is just increased). Additional evidence that the EG are 'activated' could be provided by looking at the increase in Draper intensity (as reported by Doherty et al. and MacDonald et al. that the authors cite), not only in the olfactory system, but also in the visual system and in the MB. It would also be informative if the authors can look at morphology of the EG in the visual system and MB to convincingly that the data shown in Fig4 is relevant to EG 'activation'.

Major Concern #3: In Fig3, there is no clear explanation why they focus on the ON transients and ignore the OFF transients, and also why the difference in the depolarization is not quantified in Fig4.

Major Concern #4: While the authors claim that mz709-GAL4 is a EG specific driver, do the authors know that this is indeed true in the tissues and stages that are studied here? The Ito et al,. paper that is cited in the METHOD section has only looked at the expression of this reporter in embryonic and larval stages. The authors need to that the authors should validate their findings with an additional EG specific driver and/or provide additional data that mz709-GAL4 is indeed specific to EG in the adult fly brain and eye. If mz709-GAL4 is expressed in other cell-types, the interpretation of many of the data in this paper becomes quite questionable. I believe the data in Fig3B is suggesting that mz709-GAL4 is indeed specific to glia cells and not expressed in neurons, but whether this driver is truly specific to EG (and not in other glial types), especially in the visual system (including the lamina as well as in the eye), is not obvious.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study investigates the impact of Pink1 loss on glial function and neuronal health in a Drosophila model, highlighting the role of mitochondria-organelle contacts and key genes such as Ccz1, Vps13, Mon1, and Rab7. The work provides insights into cellular processes underlying neurodegenerative diseases, with a focus on glia-neuron interactions. While the findings are promising, the study lacks critical controls, detailed mechanistic evidence, and explanatory figures to strengthen its claims.

Strengths:

(1) The study addresses an important topic in neuroscience, exploring the mechanisms of Pink1 loss, which has implications for Parkinson's disease and neurodegeneration.

(2) The focus on mitochondria-organelle contacts and their regulation by Rab7-mediated pathways is novel and provides a potential mechanism for neuronal dysfunction.

(3) The identification of key genes (Ccz1, Vps13, Mon1, Rab7) and their potential roles in Pink1-related pathways adds valuable knowledge to the field.

(4) The manuscript uses a combination of genetic tools, Drosophila models, and functional assays to approach the problem from multiple angles.

Weaknesses:

(1) Specificity of Mz-Gal4: The study lacks validation of Mz-Gal4 specificity, as it may also drive expression in a few neurons or other types of glia. Additional control experiments using nls-GFP with Elav, Repo, or Draper antibody staining or alternative glial drivers would be helpful.

We have addressed this issue of Gal4 driver specificity based on new experiments in the revised manuscript.

(2) DLG staining is central to the story but is not well-supported by high-resolution Z-stack imaging, which should be included in the supplementary figures.

We have included these in the supplement.

(3) The manuscript does not confirm whether the candidate RNAi (Ccz1, Vps13, Mon1, Rab7) directly influence Rab7-mediated membrane trafficking or mitochondria-lysosome contacts in Pink1 mutants.

This is indeed the case. These more mechanistic experiments were not yet performed.

(4) Using ERG as a readout for EG effects in the antenna is not a direct or appropriate assay. Alternative functional assays relevant to antenna glia should be considered.

We made the assumption that ensheating glial function is conserved across brain regions and now make this explicit in the reworded manuscript.

(5) A graphical explanation of the interactions and functions of the candidate genes in Pink1 KO mutants is missing. This would greatly enhance the manuscript's clarity.

We have included such a scheme in the new manuscript.

(6) The study lacks details on sample sizes, effect sizes, and reproducibility, which are necessary for robust conclusions.

We have included these essential data in the reworked document.

(7) There are repeated words on page 3 ("olfactory Olfactory Receptor Neurons") and a lack of explanation in Figure 3C regarding the most up-regulated and down-regulated genes and the significance of large red dots.

We have included the requested information.

Reviewer #2 (Public review):

Summary:

This study proposes a novel role for ensheathing glia (EG) in a Pink1-model of Parkinson's disease and shows that this cell population exibits the highest number of DEG in a pre-symptomatic stage. In the olfactory system, there seems to be morphological changes in this cell-type that resembles an 'activated' state and the authors further show that the neuronal loss of Pink1 is responsible for this defect. The authors go on to show that manipulation of Pink1 in EG also leads to some defects in the visual system and in the dopaminergic neurons (DAN) that innervate the mushroom body (MB), and performed a screen based on the 'on-transient' defect of the ERG to identify potential genes that may modulate the function of EG in synaptic regulation. They focus on several genes related to Rab7/Vps13, and performed some additional experiments in the visual system and MB to propose the role of vesicle/lipid trafficking in EG as a important factor for PD pathogenesis.

Strengths:

The study proposes functional and mechanistic connections between several genes that have been linked to PD (PINK1, VPS13A/C). I feel that the data presented in Figure 1 and Fig3A-C are performed with rigor and are convincing/novel. The selection of Drosophila to study the questions is also a strength and the lab has extensive experiences in this field and model organism.

Weaknesses:

There is one fundamental concern I have with the genetic experiments performed in this paper (especially in Fig 3D and Fig4, see major issue #1), and I feel that there is a bit of a disconnect between the EG 'activation' phenotype the author show in the olfactory system and the other two neuronal systems (visual system, MB DAN) that the authors investigate see major issue #2). Also, there are quite a bit of information that is not provided in the manuscript (see major issues #3 and #4), which makes me difficult to judge the rigor and interpretation of several experiments.

Major Concern #1: A number of lines used in this study are referred to as "RNAi" lines but when I look at the actual genotypes of reagents listed in the table in the METHODS section, many are actually NOT RNAi lines. Quite a few lines, including lines that the authors use as RNAi against Ccz1, Rab7 and Mon1, are gRNA lines for the TKO (TRiP-CRISPR knockout) system. While these reagents can theoretically knock-out these genes in somatic cells if used in combination with UAS-Cas9, there is no mention that UAS-Cas9 was used in this work throughout the manuscript. Hence, when these lines are just crossed to GAL4 with or without the Pink1 mutant, they shouldn't be having any effects. Similarly, the strongest hit from their screen was a TOE (TRiP-CRISPR Over Expression) gRNA against PIG-A, which could allow overexpression of PIG-A if there is a UAS-dCas9::VP64. However, I also do not see any mention that such activator was introduced into the crossing scheme. Considering that 3 of the 4 'hits' from their screen are not RNAi lines, I am quite skeptical of the study. Similarly, except for Vps13, all reagents used in Fig4 are TKO gRNA lines. Therefore, if this experiment was conducted without an UAS-Cas9, most of the data shown here are problematic. Also, note that several of the 'RNAi' lines listed in the Table in the METHODS section are actually MiMIC alleles. While some MiMIC lines could function as strong LOF alleles (if they are inserted in the exon or in an intron of the gene in the same orientation as the gene), some of the lines are not expected to affect gene function (e.g. FASN2 and CG17712, MiMICs are in introns and face the opposite orientation). Hence, the rationale of including these reagents in the screen doesn't make much sense. The description of the modifier screen should be much more detailed in the RESULTS and METHODS section and if the UAS-Cas9/dCas9::VP64 transgenes were not introduced when the TKO/TOE reagents were utilized, what can be concluded?

In addition, for the 4 genes that the authors further study in Fig4, there are many other reagents that the authors can use, including mutant alleles, previously characterized RNAi lines (e.g. Vps13) and dominant negative/constitute active lines (e.g. especially for Rab7). The authors should validate their results with independent reagents to really convincingly show that the same conclusions can be drawn for the Vps13/Rab7 related genes since this is the key takeaway message of this paper.

Also, they do not show whether the manipulation of these genes in a wild-type background (they only show what happens in Pink1 mutants) affect ERG and MB DAN synapse morphology. If these manipulations alone dramatically affect these phenotypes, it would be very difficult to interpret their data.

We sincerely thank the reviewer for spotting this major oversight regarding the use of the TKO (TRiP-CRISPR knockout) and TOE (TRiP-CRISPR Over Expression) systems and the MiMIC alleles. As the reviewer pointed out, these lines were not used as intended, therefore our results and conclusions regarding the genetic interactions between Pink1 and several genes (PIG-A, Rab7, Ccz1, CG10646, Mon1, FASN2, CG17712), are incorrect and based on a technical mistake. These results were removed from the manuscript. While our mistake compromises the data regarding PIG-A, Rab7, Ccz1, CG10646, Mon1, FASN2, CG17712, it does not affect the results and conclusions for most of the genes of the screening and for Vps13 where we did use RNAi lines.

Also, in the reworked manuscript, we provide additional evidence that modulation of vesicle trafficking proteins involved in mitochondria–endoplasmic reticulum (ER) membrane interactions, such as Vps13 and Vps35, influences neuronal function and rescues Pink1 mutant phenotypes when selectively downregulated in EG.

Major Concern #2: In Figure 1, the authors show some morphological evidence that EG are 'activated' in Pink1 mutants, but whether the same phenomenon occurs in the visual system and in the MB is not shown. Since all of the studies in Fig3D and Fig4 are done in the visual system and MB, it is not clear whether the visual system and MB phenotypes are related to 'activation' of EG.

Also, in the RNA-seq data in Fig1A and Fig3C, is there any molecular evidence that EG are indeed 'activated'? The only evidence that the authors show to state that EG are 'activated' in young Pink1 null animals is based on increased CD8::GFP staining in the olfactory system.

The authors cannot draw a strong conclusion that indeed EG are 'activated' based on these data (e.g. perhaps the expression level of CD8::GFP is just increased). Additional evidence that the EG are 'activated' could be provided by looking at the increase in Draper intensity (as reported by Doherty et al. and MacDonald et al. that the authors cite), not only in the olfactory system, but also in the visual system and in the MB. It would also be informative if the authors can look at morphology of the EG in the visual system and MB to convincingly that the data shown in Fig4 is relevant to EG 'activation'.

In line with the identification of DEG across the ensheating glia cluster in our single cell sequencing (where we did not distinguish between EG of different brain regions) we made the assumption that EG-(dys) function is consistent in the Pink1 mutant and conserved across brain regions. Nonetheless, to make clear that we did not consistently analyze EG morphology in the different brain regions that we probed in functional assays, we added a note in the manuscript. Furthermore, we also toned down our conclusion that the EG in Pink1 mutants are in an activated state: we note the similarity in phenotype in Pink1 mutants and situations of neuronal damage (where EG are activated) but added that the phenotype in Pink1 mutants may also be the result of the mere upregulation of GFP expression/fluorescence.

Major Concern #3: In Fig3, there is no clear explanation why they focus on the ON transients and ignore the OFF transients, and also why the difference in the depolarization is not quantified in Fig4.

We included this explanation in the reworked manuscript: In the Drosophila ERG, the sustained depolarization primarily reflects phototransduction in photoreceptors (and is defective when photoreceptors degenerate), whereas the ON and OFF transients arise from second-order lamina neurons and are widely used as readouts of signal transfer. We wanted to assess function and focused on the ON transient because in general it provides an onset-locked, more robust readout of function (Vilinsky & Johnson, 2012).

Major Concern #4: While the authors claim that mz709-GAL4 is a EG specific driver, do the authors know that this is indeed true in the tissues and stages that are studied here? The Ito et al,. paper that is cited in the METHOD section has only looked at the expression of this reporter in embryonic and larval stages. The authors need to that the authors should validate their findings with an additional EG specific driver and/or provide additional data that mz709-GAL4 is indeed specific to EG in the adult fly brain and eye. If mz709-GAL4 is expressed in other cell-types, the interpretation of many of the data in this paper becomes quite questionable. I believe the data in Fig3B is suggesting that mz709-GAL4 is indeed specific to glia cells and not expressed in neurons, but whether this driver is truly specific to EG (and not in other glial types), especially in the visual system (including the lamina as well as in the eye), is not obvious.

We labelled animals that express UAS-HisTag-eGFP (used also in our paper) under control of MZ709-Gal4 with anti-Elav (a neuronal marker) and find no significant overlap (see below “recommendation for authors”), consistent with MZ709-Gal4 not driving expression in neurons. This is consistent with previous published work: Indeed, MZ709-Gal4 has been amply used in adult flies and shown to be ensheating glia-specific (Doherty et al., 2009; Li et al., 2023; Sehgal et al.,2018). In the lamina neuropil of the Drosophila eye, MZ709-Gal4 is expressed in the marginal glia (Stenesen et al., 2019) which are neuropil-associated glia and are equivalent to generic ensheathing glia (Kremer et al., 2017). MZ709-Gal4 is also expressed also in satellite glia (Stenesen et al., 2019), but these glia enwrap the cell bodies of the lamina neurons and not the neuropil where synapses reside.

Recommendations for the authors:

Reviewing Editor Comments:

We strongly encourage you to very carefully edit this manuscript. The reviewers made many probing comments that you should consider carefully.

Reviewer #1 (Recommendations for the authors):

(1) Validate the specificity of Mz-Gal4 by performing experiments with nls-GFP and Elav antibody staining to ensure there is no neuronal overlap. Additionally, consider using alternative glial-specific drivers, such as Repo-Gal4 or WG-Gal4, to confirm the findings.

We expressed HisTag-eGFP (used also in our paper) under control of MZ709-Gal4 and labelled fly brains with anti-Elav (a neuronal marker). We do not observe significant overlap between the labels indicating MZ709-Gal4 does not express Gal4 in neurons (Supplementary figure 1).

As indicated, these observations are consistent with previous published work. MZ709-Gal4 has been amply used in adult flies and shown to be ensheating glia-specific (Doherty et al., 2009; Li et al., 2023; Sehgal et al., 2018; Stahl et al., 2018). In the lamina neuropil of the Drosophila eye, MZ709-Gal4 is expressed in the marginal glia (Stenesen et al., 2019) which are neuropil-associated glia and are equivalent to generic ensheathing glia (Kremer et al., 2017). MZ709-Gal4 is also expressed also in satellite glia (Stenesen et al., 2019), but these glia enwrap the cell bodies of the lamina neurons and not the neuropil where synapses reside.

(2) Include high-resolution Z-stack imaging of DLG staining to strengthen the assessment of synaptic integrity and ensure the robustness of the conclusions. These images should be added to either the main or supplementary figures.

We included 2 supplementary figures (2 and 3) showing Z stacks that were used to delineate regions of interest at the MBs for the quantification of dopaminergic neuron afferents invasion. Our approach is identical to the one we used in Kaempf et al. 2026 (Kaempf et al., 2026).

(3) Demonstrate whether the candidate RNAi (Ccz1, Vps13, Mon1, Rab7) directly influence Rab7-mediated membrane trafficking or mitochondria-lysosome contacts in Pink1 mutants. Use an appropriate method to confirm changes in organelle contacts in response to the RNAi treatments.

Ccz1, Mon1 and Rab 7 were removed due to the technical mistake we made. We did confirm and maintain that Vps35 and Vps13 downregulation in EG rescues neuronal defects in Pink1 mutants. In the reworked manuscript we present a possible mechanism that involves the role of Vps35 and Vps13 in regulating ER-mitochondrial contacts, in line with our previous work (Valadas et al., 2018), while not ruling out possible other mechanisms.

(4) Provide an alternative functional assay or evidence to support the use of ERG as a readout for EG effects in the antenna. Consider using a more direct assay relevant to antenna glia function.

We agree that a more direct functional assay of antennal glia would be a nice addition (e.g., single-sensillum recordings or glial/ORN Ca2+ imaging). However, implementing such assays would require new experimental pipelines and substantial additional data generation that is beyond our current ability and the scope of this revision.

(5) Add a graphical illustration explaining the proposed mechanism of how Ccz1, Vps13, Mon1, and Rab7 function in Pink1 KO mutants, highlighting their interactions and roles within specific cell types.

We included a schematic of our working model in Figure 5.

(6) Clarify Figure 3C by explaining the most up-regulated and down-regulated genes and the significance of the large red dots. This will enhance the interpretability of the data.

We expanded the legend to this figure: The large red dots represent the genes that rescue Pink1KO-WS phenotype when downregulated, the dark green dots are the 50 top most deregulated genes (magnitude of deregulation) in EG in Pink1KO-WS compared to controls, while the light green dots represent whole the genes detected in our cell-type specific transcriptomic experiment.

(7) Correct repeated words on page 3 ("olfactory Olfactory Receptor Neurons") for clarity and consistency.

Of course, sorry for this.

(8) Ensure that sample sizes, effect sizes, and the number of replicates are explicitly stated for all experiments. This information is essential for evaluating the robustness and reproducibility of the findings.

We made sure we consistently added all this information in the revised manuscript.

(9) Verify and ensure that all data, reagents, and code used in the study are accessible and appropriately documented, in adherence with eLife's publishing policies.

We made sure all data, reagents and code are available and/or properly described.

By addressing these recommendations, the authors will significantly improve the clarity, rigor, and reproducibility of the manuscript.

Reviewer #2 (Recommendations for the authors):

Minor Points.

(1) All figures seem to lack titles.

We fixed this error.

(2) In the abstract, the authors say that Rab7 and Vps13 are mutated in PD patients but I couldn't find the reference/information for Rab7 (the authors do refer to papers that linked VPS13A/C variants to PD but no mention about RAB7A/B being linked to PD). Please discuss this in the paper or modify the abstract accordingly.

We removed this statement for rab7 from the paper.

(3) When referring to the human gene, Pink1 should be written as PINK1 according to the HGNC nomenclature rules.

We made this change.

(4) The authors say Vps13 has two mammalian orthologs but actually it has four (VPS13A/B/C/D). I guess two of the four is linked to PD so the authors should modify there statement to reflect this.

This is a misinterpretation of what we meant and we have clarified our intention: Drosophila possesses 3 paralogues of Vps13 - Vps13, Vps13B, and Vps13D - which we also detected in our screening (Neuman et al., 2025; Velayos-Baeza et al., 2004; Vonk et al., 2017). Among these Vps13 is most similar to human VPS13A and VPS13C (Hanna et al., 2023; McEwan & Ryan, 2022).

(5) The abbreviation 'CNS' is used in the first page of the intro but I don't see it being spelled out as "central nervous system".

We have spelled out central nervous system in the first page of the introduction.

(6) On the top of page 5, the authors state that they confirmed that the 'synaptic area of DAN show a decrease in aged (25 days) animals' but data is not shown. If they want to make a statement like this, I believe such data should be included in supplemental data. Since the phenotype in the aged animal is not relevant to this study, one could remove this statement regarding the aged animals if they prefer not to show the data.

The decreased synaptic area of DAN in 25-day old Pink1 mutants is shown in figure 2C-D of the manuscript and is consistent with data shown in (Kaempf et al., 2026).

Version 2

Abstract

Parkinson’s disease (PD) is commonly associated with the loss of dopaminergic neurons in the substantia nigra, but many other cell types are affected even before neuron loss occurs. Recent studies have linked oligodendrocytes to early stages of PD, though their precise role is still unclear. PINK1 is mutated in familial PD, and through unbiased single-cell sequencing of the entire brain of Drosophila Pink1 models, we observed significant gene deregulation in ensheathing glia (EG); cells that share functional similarities with oligodendrocytes. We found that the loss of Pink1 leads to abnormalities in EG, similar to the reactive response of EG seen upon nerve injury. Using cell-type-specific transcriptomics, we identified deregulated genes in EG as potential functional modifiers. Specifically downregulating two trafficking factors in EG, Vps35 and Vps13, also mutated in PD, was sufficient to rescue neuronal function and protect against dopaminergic synapse loss. Our findings demonstrate that Pink1 loss in neurons triggers an injury-like response in EG, and that Pink1 loss in EG in turn disrupts neuronal function. Vesicle trafficking components, which may regulate membrane interactions between organelles in EG, seem to play a role in maintaining neuronal health and ultimately preventing dopaminergic synapse loss. Our work highlights the essential role of glial support cells in the pathogenesis of PD and identifies vesicle trafficking within these cells in disease progression.

Introduction

Parkinson’s disease (PD) is characterized by motor symptoms such as bradykinesia, rigidity and tremor that are caused by the progressive loss of dopaminergic neurons in the substantia nigra (Lang & Lozano, 1998). However, most of the patients report non-motor symptoms, such as constipation, hyposmia and sleep defects, even before the onset of motor symptoms (Munhoz et al., 2015). This suggests that PD is a progressive neurodegenerative disease that initiates many years before the diagnosis and involves different neuronal systems, multiple anatomical areas and different cell types. While therapies symptomatically improve motor functioning temporarily by restoring dopaminergic tone, they do not effectively prevent neurodegeneration and disease progression. This highlights the necessity to identify cell-types and biological mechanisms that act at the earliest stages of disease.

With the advent of single-cell sequencing applied to PD and control brains in combination with evidence from genome-wide association studies (GWAS), it is now possible to identify which cells and mechanisms are at play in PD models that recapitulate early stages of PD (Kaempf et al., 2026; Pech et al., 2025). Among these, one study reported that genes related to GWAS loci are enriched for oligodendrocyte-specific gene expression (Bryois et al., 2020). Interestingly, differential gene expression in post-mortem brains with Braak scores of 1-2, 3-4, and 5-6 and controls, suggested a role for oligodendrocytes at an early stage of PD prior to the overt loss of dopaminergic neurons (Bryois et al., 2020). Additionally, two independent single-cell sequencing studies, one of the human substantia nigra and one of the midbrain, both associated PD genetic risk with oligodendrocyte-specific expression patterns (Agarwal et al., 2020; Smajic et al., 2022). While these studies started to link dysfunction of oligodendrocytes to the early stages of PD, they did not address their role in the pathophysiology of the disease.

Here, we describe a Drosophila model to start assessing how neuron-glia cross-talk contributes to PD. Drosophila has been successfully used to investigate cellular and molecular dysfunction preceding the onset of age-related symptoms, including dopaminergic neuron-dependent motor symptoms (Kaempf et al., 2026; Pech et al., 2025; Valadas et al., 2018). Furthermore, Drosophila glial cells display several anatomical and functional features that show remarkable similarity to their mammalian counterparts (Freeman & Doherty, 2006; Kremer et al., 2017). While the small size of the Drosophila brain does not necessitate elaborate myelination, ensheathing glia (EG) in this species share not only anatomical features with oligodendrocytes, such as the ability to wrap around nerve tracts in the Central Nervous System (CNS) (Kremer et al., 2017; Yildirim et al., 2019), but also important molecular and functional features, such as providing metabolic support and regulating neuronal activity (Delgado et al., 2018; Otto et al., 2018).

In a brain-wide single-cell sequencing experiment, we identified EG as the most deregulated cell type in a young (pre-motor) Pink1 loss-of-function Drosophila model. Using cell-specific labeling, immunohistochemistry, and electrophysiological recordings, we correlated the transcriptional deregulation in EG with defects that appear similar to those seen upon nerve injury. Additionally, we find that healthy EG are fundamental to support dopaminergic synapse integrity in Pink1 mutants. Finally, using cell-type-specific transcriptomic, high-throughput screening and immunohistochemistry, we identified vesicle trafficking as a genetic modifier and showed that Vps13 and Vps35 are EG-expressed regulators that maintain dopaminergic synapse integrity. Our study shows that neuron-ensheathing glia crosstalk is already disrupted at a young age by the loss of Pink1 in neurons, but also in EG, and that EG are fundamental to support dopaminergic synapse-integrity, suggesting a role for oligodendrocytes in the progression of PD.

Results

Ensheathing glia in Pink1 loss-of-function flies show cell non-autonomous defects

We previously created a whole-brain single-cell RNAseq dataset from different Drosophila PD knock-in models. This dataset was generated from young 5-day-old flies to capture ‘early’ changes (Pech et al., 2025). We re-analyzed the data from Pink1P399L knock-in loss-of-function mutant flies and isogenic controls and performed differential gene expression (DEG) analysis for each cell type using DESeq2 (Figure 1A). Glial cell subtypes show the highest degree of deregulation; particularly, ensheathing glia are strongly affected (Figure 1A). While flies do not myelinate neurons, ensheathing glia (EG) serve similar supporting functions as oligodendrocytes (Otto et al., 2018; Yildirim et al., 2019). These results suggest that ensheathing glia are deregulated in young Pink1P399L mutants (and also Pink1KO-WS, see below) at an age prior to dopaminergic neuron-dependent motor defects (Kaempf et al., 2026; Pech et al., 2025).

EG are affected non-cell autonomously by Pink1 loss-of-function (with supplementary fig 1)

(A) tSNE of the cells of Pink1P399L knock-in mutants (5-day-old). Cell types are labeled with colors indicating the number of deregulated genes compared to control. EG are encircled and labeled. (B-B”) Maximum intensity projections of confocal images of fly brains (5 ± 1-day-old) stained with anti-GFP (Green) and anti-Brp (Magenta), where anti-GFP marks EG and anti-Brp marks presynaptic sites of the antennal lobes in flies where CD8GFP is expressed via the EG driver MZ709-Gal4. Scale bar: 20 µm (B’) Maximum intensity projection of confocal images of controls vs. controls 24 hours after ORN-severing (injury). (B”) Maximum intensity projection of confocal images of Pink1KO-WS vs Pink1KO-WS 24 hours after ORNs severing (injury). (C) Quantification of GFP intensity within the glomeruli of the antennal lobe area in 5 ± 1-day-old flies (as in B) relative to controls. ANOVA with Dunnet’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η =0.23. Bars: mean ± SD; points are individual animals N≥13 per genotype, 4 replicates. (D-D”) Maximum intensity projection of confocal images of fly brains (5 ± 1-day-old) stained with anti-GFP (Green) and anti-Brp (Magenta), where anti-GFP marks EG and anti-Brp marks presynaptic sites of the antennal lobes in flies where CD8GFP is expressed via the EG driver MZ709-Gal4. Scale bar: 20 µm. (D’) Maximum intensity projection of confocal images of control (w1118) and Pink1KO-WS animals (D”) Maximum intensity projection of confocal images of animals with Pink1 downregulation in EG and Pink1KO-WS with Pink1 rescued in EG. (E) Quantification of GFP intensity within the glomeruli of the antennal lobe area in 5-day-old flies (as in D) relative to the control. ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η =0.36 Bars: mean ± SD; points are individual animals N≥10 per genotype, 4 replicates.

When neurons are damaged, activated EG invade the neuropil (Doherty et al., 2009; MacDonald et al., 2006). For example, when fly antennal lobes are removed, severing the Olfactory Receptor Neurons (ORNs), EG invade the antennal lobe neuropil (MacDonald et al., 2006), potentially as a protective response (Figure 1B’-C). To test if EG invaded the neuropil in Pink1 mutant flies, we expressed a transmembrane fluorescent protein (UAS-CD8GFP) under the control of a promoter specific to EG (MZ709-Gal4) (Doherty et al., 2009, Supplementary figure 1) in control and Pink1KO flies. Samples were labeled with anti-GFP and the pre-synaptic protein Bruchpilot (NC82). Similar to ORN-severed controls, GFP signal, lining the membrane of EG, is visible inside the antennal neuropil in Pink1KO, and severing the ORN in Pink1KO does not further exacerbate this phenotype (Figure 1B”-C). Even though we cannot exclude that in Pink1 mutants the GFP signal is merely upregulated, Pink1 loss and neuron-severing that triggers EG activation seem to have a similar phenotype.

In order to determine whether this phenotype is cell-autonomous, we generated cell-type-specific Pink1 perturbations. We either downregulated Pink1 using a previously well-validated RNAi line specifically in EG (EG-specific Pink1 loss-of-function – Figure 1D’), or we re-expressed wild-type Pink1 specifically in EG in Pink1KO-WS flies (all cells, but the EG, loss-of-function - Figure 1D”). When Pink1 is knocked-down in EG (but present in other cells, including neurons), no GFP signal is detected in the neuropil (Figure 1D’’-1E). However, when Pink1 is not present in neurons and expressed in EG, the GFP is present inside the antennal neuropil (Figure 1D”-E). This suggests that Pink1-defective neurons activate the EG in a cell-nonautonomous fashion to invade, and potentially protect, the neuropil.

Ensheathing glia function supports synaptic integrity

Our finding that Pink1KO-WS in neurons elicits a cell non-autonomous response in EG suggested that EG might, in turn, also functionally modulate neuronal integrity in mutant flies. To test this, we made use of a readily accessible model circuit in the fly visual system with an easy electrophysiological read-out, electroretinograms (ERG). Notably, whereas our functional readout here interrogates neuronal activity in the visual system, our analysis of EG morphology was performed in the antennal lobe, a distinct brain region. We therefore make the explicit assumption that EG perform comparable functions across fly brain regions. This assumption is supported by our single-cell sequencing dataset, in which we did not detect region-specific segregation of EG into distinct clusters that could be attributed to different brain areas. The ERG waveforms are susceptible to changes in intracellular signaling, neuronal function, and synaptic transmission, and have been well characterized and amply used to assess neuronal and cellular function (Hardie & Raghu, 2001; Praschberger et al., 2023; Soukup et al., 2016; C. F. Wu & Wong, 1977). We exploited this method to understand the role of Pink1 in EG on neuronal function and synaptic transmission in young flies.

Comparing control with young Pink1KO-WS flies does not show a difference in the “depolarization-response”, indicating that -as expected-there is no neurodegeneration occurring in the photoreceptors at this stage (Figure 2A). However, when comparing the ON peak – which represents synaptic transmission from the photoreceptors and within the underlying neuronal circuit, and is a more robust readout of this than the OFF peak (Vilinsky & Johnson, 2012) – we detect a significant reduction in mutant compared to control animals, suggesting that in Pink1KO-WS flies, synaptic communication in this circuit is partially impaired, already at 5-day-old (Figure 2A-B).

Pink1 in EG is necessary to support synaptic integrity (with Supplementary figure 2)

(A) Representative ERG traces of indicated genotypes: ON peak is highlighted by the arrow. (B) Normalized ON peak amplitude of flies (5 ± 1-day-old). ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η =0.55 Bars: mean ± SD; points are individual animals N≥11 per genotype, 3 replicates. (C-C”) (C) Maximum intensity projection of confocal images of control and Pink1KO-WS in Mushroom Bodies (MB) of aged flies (22 ± 2-day-old), stained with anti-TH (cyan) and anti-DLG (magenta) antibodies - DLG is used to mark post-synaptic sites of MB. The black and white image is the middle Z-plane within the region of interest of the MB (ROI, yellow), which is used to represent the thresholded TH area (white). Scale bar: 20 µm. (C’) Maximum intensity projection of confocal images of w1118 with Pink1 downregulation in EG. (C”) Maximum intensity projection of confocal images of Pink1KO-WS with Pink1 rescued in EG. (D) Quantification of the dopaminergic synaptic area at MB neuropil in aged flies (22 ± 2-day-old). ANOVA with Tukey’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η = 0.38. Bars: mean ± SD; points are individual animals N≥12 per genotype, 3 replicates.

In order to discern the role of EG in this defective ERG response, we (1) downregulated Pink1 specifically in EG and (2) we expressed Pink1 specifically in the EG of Pink1KO-WS. When Pink1 is knocked down specifically in EG, we observe the same reduction in ON peak response as in the Pink1KO-WS flies. This indicates that EG integrity is necessary for efficient neurotransmission in this circuit. Conversely, the ON peak defect of Pink1KO-WS mutants is rescued when Pink1 is reintroduced in EG only. This suggests that activated EG are able to, at least partially, buffer the detrimental effects of Pink1 deficiency in neurons (Figure 2A-B).

To also test EG function in relation to PD-relevant dopaminergic (DA) synapses, we evaluated the loss of dopaminergic neuron afferents. In a previous paper from our laboratory, we have shown that the loss of Pink1 function causes a progressive loss of protocerebral anterior medial dopaminergic neuron (PAM DAN) afferents in the Mushroom Body (MB) neuropil in 25-day-old animals, but not when they are 5-day-old (Kaempf et al., 2026). Indeed, we confirm that 25-day-old animals show a decrease in the dopaminergic synaptic area in MB lobes (Figure 2 C-D). Hence, while ensheathing glia are already activated at a young age in Pink1KO-WS, DA synapses are structurally still intact (at 5 days of age) and only deteriorate in older flies. We then knocked down Pink1 specifically in EG and found a similar strong decrease in DA synaptic area (Figure 2C’-D). Conversely, when Pink1 is reintroduced specifically in EG of Pink1KO-WS animals, DA synapse-loss is significantly rescued (Figure 2C”-D and Supplementary figure 2). Together, these results indicate that Pink1 in EG is necessary to protect dopaminergic neurons from synapse loss.

EG cell type-specific transcriptomics reveals modifiers of neuronal dysfunction

To identify pathways in EG that are deregulated by Pink1 loss-of-function, we optimized a method to isolate EGs from the rest of the brain, enabling us to perform much higher-sensitivity transcriptomics than what we achieved with our 10x droplet-based single-cell sequencing (Pech et al., 2025). We expressed a fluorescently-tagged histone (UAS-HisTag-eGFP) in EG (GMR-56-Gal4) and used Fluorescence-activated Cell Sorting (FACS) to isolate EG. We sorted 300 cells per technical replicate and performed Bulk-RNA sequencing using a SMART-seq2 protocol (Figure 3A). The sequencing results of EG-sorted cells and of neuron-sorted cells (nSyb-Gal4 instead of GMR-56-Gal4) show strong enrichment of EG cell markers in the former and strong enrichment of neuronal markers in the latter (Figure 3B). We then compared the transcriptomic profile of EG in Pink1KO-WS to that of control using differential expression analysis and found 617 deregulated genes in EG (Appendix 1, padj<0.05) (Figure 3C). To understand whether the deregulated genes were enriched for a specific pathway or associated with a cellular component compared to non-significantly deregulated genes, we performed Gene Ontology analysis of the deregulated genes in Pink1KO-WS flies. Our analysis resulted in generic terms and very diverse pathways that did not allow us to draw further conclusions.

Cell type-specific transcriptomics reveals modifiers of neuronal dysfunction

(A)Scheme of cell-type specific transcriptomics (created using BioRender.com). (B)Scaled gene expression of representative genes for EG and neurons after sorting EG or neurons using the protocol described in (A) (R Core Team, 2022). N=2, 2 replicates. (C)Differentially expressed genes in EG in Pink1KO-WS compared to control flies, plotted according to their Log2 Fold change and the -Log10 of the adjusted p-value. Intercept in red (-Log10 adjusted p-value = 4.31); light green dots are all detected genes, dark green are the 50 most deregulated genes, and the pink dot is from a gene positive in the genetic screen (D). *Two data points are outside the boundaries of the plot. To determine the transcriptomic profile of each genotype, a N=3 was used, and 3 replicates were performed. (D)ERG ON peak value differences to Pink1KO-WS flies (5 ± 1-day-old) of control (red) and of Pink1KO-WS flies with DEGs downregulated, or upregulated, specifically in EG (5 ± 1 days old). ON peak values are expressed as the difference to Pink1KO-WS. ANOVA with Dunnett’s test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η = 0.44. Bars: mean ± SD; points are individual animals N≥3 per genotype. *One data point is outside the boundaries of the plot.

To test if the genes deregulated in EG are modifiers of the neuronal pink1KO-WS phenotypes, we downregulated and upregulated (when possible) the 50 most deregulated genes (Figure 3C), for which we could find RNAi fly lines, specifically in the EG of Pink1KO-WS flies and recorded ERGs in 5-day-old animals. Knock down of one of the genes (CG17660) resulted in a rescue of the ON transient defects in Pink1KO-WS (suppressors at p<0.05), and a few exacerbated the defect (Figure 3D). Interestingly, the human orthologs of CG17660, TMEM87A/B, have an established role in endosomal sorting (Gaudet et al., 2011; Hirata et al., 2015). Hence, our genetic screening suggests that Pink1KO-WS phenotypes might be suppressed by downregulating genes involved in vesicle trafficking in EG.

PD causative vesicle trafficking gene downregulation in EG rescues synaptic dopaminergic neurons afferent loss

Because our ERG-based modifier screen for the Pink1KO-WS phenotype identified a gene with a known role in vesicle trafficking, we next asked whether Pink1 in EG also genetically interacts with other PD-causative genes that function in vesicle trafficking, specifically vesicle trafficking components affecting membrane interactions between mitochondria and the endoplasmic reticulum (ER), like Vps35 and Vps13 (Brickner & Fuller, 1997; Lesage et al., 2016; Vilariño-Güell et al., 2011; Zimprich et al., 2011). We find that the knock down of each of these genes in EG also rescues the ON transient defect of Pink1KO-WS mutants (Figure 4A-B).

Vps35 and Vps13 downregulation in EG rescues synaptic deficits in Pink1KO-WS flies (with supplementary figure 3)

(A)Representative ERG traces of control,Pink1KO-WS flies and Pink1KO-WS flies with Vps35 or Vps13 downregulated in EG. (B)Quantification of the normalized ON peak response of flies with the genotypes in (A) (5 ±1-day-old). ANOVA with Dunnett’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η = 0.48. Bars: mean ± SD; points are individual animals N≥10 per genotype, 3 replicates. (C)Maximum intensity projection of confocal images of Mushroom Bodies (MB) of aged flies (22 ± 2-day-old) of control, Pink1KO-WS flies and Pink1KO-WS flies with Vps13 downregulated in EG, labeled with anti-TH (cyan) and anti-DLG (magenta), DLG is used to mark the MB neuropil. The black and white image is the middle Z-plane within the region of interest of the MB (ROI, yellow), which is used to represent the thresholded TH area (white). Scale bar: 20 µm. (D)Quantification of the dopaminergic synaptic area within MB of aged flies (22 ± 2-day-old). ANOVA with Dunnett’s multiple comparison test, * is p<0.05, ** is p<0.01, *** is p<0.001. Effect size: η =0.23. Bars: mean ± SD; points are individual animals n≥22 per genotype, 5 replicates.

To determine if manipulation of this pathway in EG can also rescue the synaptic innervation defect of PAM DAN onto the mushroom bodies in old (20-25-day-old) Pink1KO-WS mutants, we expressed RNAi to Vps13 in EG using MZ709-Gal4. Anti-TH labeling (marking DAN synapses) in the MB area (marked by anti-DLG) is significantly reduced in aged Pink1KO-WS, but is rescued to control levels when Vps13 is knocked down in EG (Figure 4C-D, Supplementary figure 3). These results suggest that manipulation of vesicle trafficking components affecting membrane interactions between mitochondria and the ER in EG regulate glia-neuron crosstalk to maintain synaptic integrity of dopaminergic neuron synapses in the fly brain of Pink1 mutants.

Discussion

In this work, we provide evidence for early, non-cell-autonomous activation of EG in a PD-relevant Drosophila Pink1 model. This occurs already in young Pink1-mutant flies when neuronal defects in dopaminergic neurons are not yet measurable. The EG activation and invasion phenotype appears as a protective response, as Pink1-deficient neurons seem to show defects in EG that mimic the response to nerve injury (Doherty et al., 2009; MacDonald et al., 2006). In addition to this, we show that there is also a non-autonomous role of EG to support Pink1-deficient neurons. When we manipulate the expression of membrane-lipid trafficking genes, the homologs of the established PD genes VPS13C and VPS35 (Lesage et al., 2016; Vilariño-Güell et al., 2011; Zimprich et al., 2011), specifically in EG, we rescue the dysfunction of Pink1 mutant neuron defects (ERG and PAM DAN synapse loss). Our work shows the importance of neuron-glia cross-talk in the context of Pink1-deficiency. We suggest (1) there is early EG activation secondary to Pink1-loss-induced neuronal impairments; (2) there are specific lipid- and membrane trafficking problems caused by Pink1 loss in EG that center on mitochondria/ER contacts and (3) there is a convergence of Parkinson-relevant genetic factors that act in EG to maintain neuronal function.

The loss of Pink1 function in the context of Parkinson’s disease has been amply linked to the regulation of mitochondrial health. Pink1 maintains the integrity of the electron transport chain by phosphorylating NDUFA10 (Morais et al., 2009, 2014; Pogson et al., 2014) and when mitochondria are damaged, it phosphorylates Parkin and Ubiquitin to facilitate mitophagy (Kane et al., 2014; Narendra et al., 2010; Narendra & Youle, 2024). The regulation of mitophagy is complex and requires mitochondrial rearrangements controlled by mitochondria-organelle contacts (ER and lysosomes) (Wong et al., 2018, 2019; Yamano et al., 2018). Such contacts mediate inter-organellar lipid exchange as well as facilitate organellar fusion (Kumar et al., 2018; Valadas et al., 2018). In Pink1 mutants, these contacts appear to be increased in number, and this causes cellular defects (Grossmann et al., 2023; Valadas et al., 2018).

We found that in EG, Vps35 functionally interacts with Pink1-induced neuronal dysfunction: its genetic manipulation in EG rescues cell non-autonomous neuronal dysfunction. Loss of Vps35 itself only in EG is sufficient to rescue neuronal Pink1 phenotypes. Pink1 loss-of-function models show increased numbers of mitochondria-ER contact sites (Valadas et al., 2018), affecting mitochondrial calcium levels (Barazzuol et al., 2020; De Brito & Scorrano, 2008; Ham et al., 2023) and dysregulating lipidic ER composition (Valadas et al., 2018) (Figure 5A). Indeed, Pink1 is necessary for ubiquitination and degradation of Mitofusin (Poole et al., 2010), which is fundamental for the mitochondria-ER tethering (De Brito & Scorrano, 2008). Interestingly, Vps35 deficiency promotes untethering of mitochondria-ER contact sites by increasing the mitochondrial levels of MUL1, which is necessary for the ubiquitination and degradation of Mitofusin (Puri et al., 2019; Tang et al., 2015; Yun et al., 2014). Hence, conditions that affect mitochondria-ER contact site regulation in EG, by modifying an endosomal trafficking regulator such as Vps35, are expected to rescue Pink1 neuronal dysfunction, possibly by restoring mitochondrial calcium levels and the lipid composition of the ER (Figure 5B); but further work is needed to sort these mechanistic elements in a cell-specific manner.

Modulation of ER–mitochondria contact sites and lipid transfer in ensheathing glia (EG) rescues Pink1-dependent neuronal dysfunction.

Schematic representation of the suggested model (created using BioRender.com). (A)Loss of Pink1 leads to an abnormal increase in endoplasmic reticulum (ER)–mitochondria contact sites (represented by blue thick lines), resulting in enhanced ER-to-mitochondria lipid transfer and dysregulation of ER lipid composition (represented by yellow lipids). Increased organelle membrane contacts and lipid flux in EG contribute to neuronal dysfunction in a non–cell autonomous manner. (B)Genetic downregulation of ER–mitochondria contact and lipid transfer regulators in EG rescues Pink1-induced neuronal phenotypes through two convergent mechanisms. Reduction of Vps35 decreases the number of ER–mitochondria contact sites, likely via MUL1-mediated Mitofusin (Mfn) turnover, leading to normalization of calcium and lipid homeostasis. In parallel, downregulation of Vps13, a lipid transfer facilitator at organelle contact sites, limits ER-to-mitochondria lipid transfer capacity, counteracting the excessive lipid flux induced by Pink1 loss. Both interventions restore organelle homeostasis in EG and result in rescue of neuronal dysfunction through a non–cell autonomous mechanism.

Our observation that downregulation of Vps13 in EG also rescues Pink1 is in further support of a role for ER-mitochondria contact site regulation. Vps13 has two mammalian homologs, VPS13A and VPS13C (Velayos-Baeza et al., 2004; Vonk et al., 2017; Vrijsen et al., 2022), which are also mutated in familial Parkinsonism/neurodegenerative disease (Lesage et al., 2016). VPS13A tethers ER to mitochondria and lipid droplets, and VPS13C tethers ER to late endosomes, lysosomes and lipid droplets (Kumar et al., 2018; Muñoz-Braceras et al., 2019; Vrijsen et al., 2022). The function of these proteins is to facilitate lipid transfer between organelles, a function we previously showed to be increased in Pink1 mutants (Valadas et al., 2018). Hence, the downregulation of Vps13 would counteract the effects of excessive organelle membrane contact formation and dysregulated lipid homeostasis in Pink1 mutants (Figure 5B).

Lipid homeostasis in glial cells is crucial for supporting neuronal function by providing energy, protecting against oxidative stress, and regulating synaptic health. In oligodendrocytes, maintaining myelin sheaths is critical (Ettle et al., 2016; Lappe-Siefke et al., 2003). While fly EG do not generate such structures, they do wrap processes around neuropil regions, also requiring lipid membrane production (Kremer et al., 2017; Pogodalla et al., 2021). Through lipid metabolism, glia also supply neurons with essential metabolites, cholesterol, and signaling molecules that aid in membrane integrity, synaptic remodeling, and neuroprotection (Delgado et al., 2018; Otto et al., 2018; Saab et al., 2016; Suárez-Pozos et al., 2020). Further work is now required to define which functions, supported by organelle contact sites in glial cells, drive cell-nonautonomous protective mechanisms in neurons. Our work in flies indicates these glial defects precede dopaminergic synapse loss, and, importantly, that rescuing the glial defects helps to prevent dopaminergic problems later in life.

Methods

Resource availability

Lead Contact

Further information and requests for resources and reagents should be directed to the lead contact, Patrik Verstreken (patrik.verstreken@kuleuven.be).

Materials Availability

Data, code, Drosophila models, and reagents are available upon request.

Fly stocks

The fruit flies were maintained in an incubator at 25°C under a 12h:12h light-dark cycle and provided with a standard diet consisting of corn meal and molasses. For experiments, only male flies were used. Flies were raised in parallel on the same batch of food, which was exchanged every three to four days and kept in only male population of similar density. Flies were aged to 5 ± 1-day-old, and 22 ± 2-day-old, as indicated for immunohistochemistry and ERGs, respectively.

To create a set of Drosophila models for Parkinson’s disease, CRISPR/Cas9-based gene editing was utilized, as outlined in (Kaempf et al., 2026; Pech et al., 2025). In brief, each knock-out line contained an attP-flanked w+ reporter cassette that replaced the first shared exon across different isoforms of the targeted gene. For this study, we used knockout flies for Pink1 from this collection (Kaempf et al., 2026). The “control” strain denotes a semi-isogenized w1118 strain that underwent backcrossing to Canton-S for 10 successive generations, resulting in a strain termed Canton-S-w1118. To generate the fly line yw; UAS-His2Av::eGFP.VK27/TM3Sb, we created the pUASTattB_His2AV-GFP plasmid, linearizing the pUAST.attB (Bischof et al., 2007) with EcoRI-BamHI. A gBlock, containing the His2AV, GS linker and eGFP sequence, was cloned into this linearized plasmid with Gibson Assembly. The plasmid was inserted into the VK27 landing site by BestGene. The sequence His2AV-GS-eGFP gblock is:

TGAATAGGGAATTGGGcAAacATGGCTGGCGGTAAAGCAGGCAAGGATTCGGGCAAGGCCAAGGCGAAGGC GGTATCGCGTTCCGCGCGCGCGGGTCTTCAGTTCCCCGTGGGTCGCATCCATCGTCATCTCAAGAGCCGCACTA CGTCACATGGACGCGTCGGAGCCACTGCAGCCGTGTACTCCGCTGCCATATTGGAATACCTGACCGCCGAGGT CCTGGAGTTGGCAGGCAACGCATCGAAGGACTTGAAAGTGAAACGTATCACTCCTCGCCACTTACAGCTCGCC ATTCGCGGAGACGAGGAGCTGGACAGCCTGATCAAGGCAACCATCGCTGGTGGCGGTGTCATTCCGCACATA CACAAGTCGCTGATCGGCAAAAAGGAGGAAACGGTGCAGGAcCCGCAGCGGAAGGGCAACGTCATTCTGTCG CAGGCCTACGGTTCAGGCGGAGGTGGCAGCGGCGGTGGCGGATCCATGGTGAGCAAGGGCGAGGAGCTGTT CACCGGGGTGGTGCCCATCCTGGTCGAGCTGGACGGCGACGTAAACGGCCACAAGTTCAGCGTGTCCGGCGA GGGCGAGGGCGATGCCACCTACGGCAAGCTGACCCTGAAGTTCATCTGCACCACCGGCAAGCTGCCCGTGCC CTGGCCCACCCTCGTGACCACCCTGACCTACGGCGTGCAGTGCTTCAGCCGCTACCCCGACCACATGAAGCAG CACGACTTCTTCAAGTCCGCCATGCCCGAAGGCTACGTCCAGGAGCGCACCATCTTCTTCAAGGACGACGGCA ACTACAAGACCCGCGCCGAGGTGAAGTTCGAGGGCGACACCCTGGTGAACCGCATCGAGCTGAAGGGCATCG ACTTCAAGGAGGACGGCAACATCCTGGGGCACAAGCTGGAGTACAACTACAACAGCCACAACGTCTATATCAT GGCCGACAAGCAGAAGAACGGCATCAAGGTGAACTTCAAGATCCGCCACAACATCGAGGACGGCAGCGTGCA GCTCGCCGACCACTACCAGCAGAACACCCCCATCGGCGACGGCCCCGTGCTGCTGCCCGACAACCACTACCTG AGCACCCAGTCCGCCCTGAGCAAAGACCCCAACGAGAAGCGCGATCACATGGTCCTGCTGGAGTTCGTGACCG CCGCCGGGATCACTCTCGGCATGGACGAGCTGTACAAATAAGGGTACCTCTAGAGGATCTT.

Flies were backcrossed for five generations into an inbred Canton-S strain harboring the w1118 mutation. The genotypes used in this study are listed in the key resource table and Supplementary table 1.

Olfactory Receptor Neurons (ORNs) axotomy

ORN bilateral axotomy was performed on 4±1 day old flies by surgical ablation of the third antennal segment. Briefly, flies were anesthetized with CO2, and then both antennae were removed with Dumont #5 forceps and returned to the vial (M. Purice, 2020; M. D. Purice et al., 2017). Flies recovered for 24 h on food at 25°C before dissection.

Immunohistochemistry and confocal imaging – invasion phenotype

Immunohistochemistry was performed on adult fly brains of 5±1-day-old, with at least two independent experiments. The brains were dissected in ice-cold PBS and fixed for 20 min in freshly prepared 3.7% paraformaldehyde (in 1x PBS, 0.3% Triton X-100 [PBX] [Sigma]) at room temperature (RT), followed by three 15-min washes in PBX at RT on a shaker. Then, the brains were incubated for 1 h in blocking solution (PBX, 10% normal goat serum [NGS]) at RT. Following blocking, the brains were incubated with primary antibodies (rabbit anti-GFP [Thermo Fisher Scientific], 1:1000, mouse anti-brp [DSHB, nc82], 1:100) in blocking solution at 4°C overnight, followed by three 15-min washes in PBX at RT. Secondary antibodies (goat anti-rabbit Alexa488, goat anti-mouse Alexa555, both at 1:1000 [Thermo Fisher Scientific]) in PBX with 10% NGS were applied overnight at 4°C. Afterwards, the brains were washed three times for 15 min in PBT at RT on a shaker and mounted with the anterior facing up in RapiClear 1.47 (SUNJin Lab). Imaging was performed using a Nikon A1R confocal microscope with a 40x (NA 1.15) water immersion lens; Z-stacks of the entire antennal lobes were acquired. The acquisition was carried out using a Galvano scanner, with a zoom factor of 1, scan speed of 0.5, and line averaging set to 2. All images were captured with a pinhole of 0.9 Airy units and a resolution of 1024 × 1024. Z-stacks (with 0,5 μm step intervals) were obtained for data acquisition, and the same imaging settings were applied across all genotypes and sessions. Image analysis was performed using Fiji (Schindelin et al., 2009). Image analysis was performed using Fiji (Schindelin et al., 2009). Each antennal lobe is composed of multiple glomeruli, and EG separate them and invade them when active (Doherty et al., 2009; Pogodalla et al., 2021). To quantify EG invasion in the antennal lobe, we calculated the invasion in every single glomerulus. To do so and be consistent with the section of the antennal lobe analyzed, anti-brp was used to identify the 3 Z-planes where DM6 and DM1 glomeruli were present (B. Wu et al., 2017). Then, to automatically detect and segment each glomerulus of the antennal lobe section, regions of interest (ROIs) were defined in the SUM projection of the 3 Z-planes by using the Fiji plugin Stardist (Weigert et al., 2019).

To quantify the EG invasion in the ROIs, anti-GFP SUM projection of the same 3 Z-planes was used. The GFP intensity was detected in each ROI, and it was normalized to the area of each ROI (GFP intensity/glomeruli area) to calculate the EG invasion in each glomerulus. Then, the normalized GFP intensity of each glomerulus of an antennal lobe was summed and normalized for the number of glomeruli detected in the antennal lobe. For each experiment, the EG antennal lobe invasion of each fly was normalized to the mean of the control. Representative images show the SUM projections of the 3 Z-planes.

Immunohistochemistry and confocal imaging – TH staining

Immunohistochemistry was performed on adult fly brains of 22±2-day-old, with at least two independent experiments. The brains were dissected in ice-cold PBS and fixed for 20 min in freshly prepared 3.7% paraformaldehyde (in 1x PBS, 0.2% Triton X-100 [PBX]) at room temperature (RT), followed by three 15-min washes in PBX at RT on a shaker. Then, the brains were incubated for 1 h in blocking solution (PBX, 10% normal goat serum [NGS]) at RT. Following blocking, the brains were incubated with primary antibodies (rabbit anti-TH [Sigma], 1:200, mouse anti-DLG [DSHB], 1:100) in blocking solution at 4°C for 1.5 to 2 days, followed by three 15-min washes in PBX at RT. Secondary antibodies (goat anti-rabbit Alexa488, goat anti-mouse Alexa555, both at 1:500 [Thermo Fisher Scientific]) in PBX with 10% NGS were applied overnight at 4°C. Afterwards, the brains were washed three times for 15 min in PBT at RT on a shaker and mounted with the anterior facing up in RapiClear 1.47 (SUNJin Lab). Imaging was performed using a Nikon A1R confocal microscope with a 20x (NA0.95) water immersion lens; Z-stacks of the entire brain were acquired. The acquisition was carried out using a Galvano scanner, with a zoom factor of 1, scan speed of 0.5, and line averaging set to 2. All images were captured with a pinhole of 2.3 Airy units and a resolution of 1024 × 1024. Z-stacks (with 3 μm step intervals) were obtained for data acquisition, and the same imaging settings were applied across all genotypes and sessions. Image analysis was performed using Fiji (Schindelin et al., 2009). To quantify dopaminergic neuron innervation of the mushroom body (MB), anti-DLG was used to identify the five Z-planes containing the synaptic region of the MB lobes. The ROI for the MB was defined in the sum projection of the five z-planes. To exclude background signal comparable to control, the area of anti-TH fluorescence was thresholded (using the default threshold for all z-planes) within the selected z-stacks. Quantification of the thresholded area within the ROI was performed in each z-plane, then summed and normalized to the MB ROI area for each brain individually (TH+ area/MB area). For each experiment, the individual TH+ area/MB area values were normalized to the mean of the control. Representative images show the maximum projection of five z-planes and the thresholded middle z-plane.

Electroretinograms (ERG)

ERGs were recorded from flies 5±1-day-old as previously described (Heisenberg, 1971; Slabbaert et al., 2016). Flies were immobilized on glass microscope slides using double-sided tape. For recordings, glass electrodes (borosilicate, 1.5 mm outer diameter) filled with 3 M NaCl were placed in the thorax as a reference and on the fly eye for recordings. Each fly was exposed to 5 cycles of 3 s of darkness, followed by a 1-s of light stimuli with LED illumination. Response to the stimuli was recorded using Axoscope 10.7 and analyzed using Clampfit 10.7 software (Molecular Devices). ERG traces were analyzed with Igor Pro 6.37 (Wave Metrics) using a custom-made macro.

Single-cell dissociation for cell-type specific transcriptomics

To test that EG could be isolated from the rest of the brain, we used two cohorts GMR56F03> UAS-His2Av::eGFP.VK27 and nSyb-Gal4> UAS-His2Av::eGFP.VK27. To identify differentially expressed genes in EG, two cohorts were used: the control w1118 expressing GMR56F03> UAS-His2Av::eGFP.VK27 and Pink1KO-WS expressing GMR56F03> UAS-His2Av::eGFP.VK27. For each genotype, ten brains of flies 5±1-day-old were collected, with most of the laminae removed, from each experimental repeat. The dissections were performed in ice-cold PBS that contained 5 μM Actinomycin D (Sigma) to inhibit changes in gene expression during tissue processing(Y. E. Wu et al., 2017). To prevent any bias caused by different experimental batches, all the genotypes were processed in parallel. To minimize variability across batches, the same reagents and procedures were used throughout the experiment. Dissections were completed within one hour. The dissection order and the assignment of genotypes to dissectors were alternated to reduce any experimenter-related variation.

For the dissociation of brain tissue, a mix of Dispase I (0.6 mg/ml) (Sigma), Collagenase I (30 mg/ml) (Thermo Fisher), Trypsin (0.5x) (Thermo Fisher), and 5 μM Actinomycin D was used. The brains were incubated for 20 min at 25°C with 1000 rpm shaking, performing four triturations at 5-min intervals. After dissociation, the cell suspension was washed with PBS containing 5 μM Actinomycin D and then filtered through a 10 μm strainer (Pluriselect), using 300 μl of PBS EDTA (Sigma) 1μM. DAPI (Sigma) (1:300.000) was added after filtration.

Fluorescence Activated Cell Sorting (FACS) for cell-type specific transcriptomics

Immediately after the dissociation, cells were sorted using a FACSAria Fusion (BD Biosciences) flow cytometer equipped with 4 lasers (405nm, 488nm, 561nm and 640nm) and a 100 µm nozzle at 20 psi. Live (DAPI-negative), GFP-positive cells were sorted into 96-well PCR plates at 300 cells per well (3 wells per genotype) using a 4-way purity mask with the FACSDiva software v9.0.1 (BD Biosciences).

Bulk transcriptomics of EG cells

Library preparations and sequencing for the Bulk RNA seq of the EG cells were performed using a modified Smart-seq2 protocol based on a previously published protocol (Picelli et al., 2013). Briefly, 3 μl of cell lysis buffer (0.1% Triton X-100, Sigma Aldrich; 1U/ul of RNase Inhibitor, RNase OUT, Thermo Fisher; 2.5 μM Oligo dT(25), IDT; and 2.5 mM dNTP, Promega) was aliquoted in triplicate to 96-well plates (4titude, Cat. No. 4TI-0960/C). Roughly 300 cells were sorted into the wells with lysis buffer and the plate was spun down at 2000xg for 1 min prior to storage at -80 °C. For the first strand synthesis, the Smart-Seq 2 plates were thawed to RT for a minute and subsequently spun down at 2000 xg for a minute. The RNA denaturation is carried out at 72°C for 10 min and then flash cooled on ice for 5 min. First strand synthesis mix (1X SuperScript II first strand synthesis buffer, Thermo Fisher; 5 mM DTT, Thermo Fisher; 100 U SuperScript II reverse transcriptase enzyme, Thermo Fisher; 10 U RNAse OUT, Thermo Fisher; 1M Betaine, Sigma Aldrich; 6 mM MgCl2, Thermo Fisher; and 1 uM TSO, IDT Technologies) is added to the cell lysis mix to a total volume of 10 μl. The First strand synthesis reaction was carried out with following program: 42°C for 90 min; 10 cycles of {50°C for 2 min; 42°C for 2 min}; 72°C for 15 min; 4°C indefinite hold. PCR amplification of the first strand product was performed by adding 15 μl of the PCR mix to the first strand product (1X KAPA HiFi HotStart Ready Mix, Roche; and 0.1 μM of IS PCR primer, IDT Technologies). 22 cycles of the following PCR program were used to amplify the first strand product: 98°C for 3 min; 22 cycles of {98°C for 20 sec; 67°C for 15 sec; 72°C for 6 min}; 72°C for 5 min; 4C indefinite hold. 20 μl of Ampure XP was added to each well and mixed. Standard Ampure XP purification was carried out as per manufacturer’s recommendation, and the cDNA library was eluted in 17.5 μl of Elution buffer. 17 μl of the eluted library was,transferred to a fresh 96-well plate (4titude, Cat. No. 4TI-0960/C). 5 μl of Tn5 tagmentation mix (1X Tagment DNA buffer, Illumina; ATM mix, Illumina, and cDNA library 1 ng) was prepared for cDNA fragmentation. Tn5 tagmentation was performed with the following program: 55°C for 10 min; 4°C indefinite hold. Tagmentation reaction was stopped by quenching the reaction with 1.25 μl of NT buffer for 5 min (Illumina). 1.25 μl of the i5 and i7 Illumina indexes were added to the plates. Finally 3.75 μl of NPM master mix was added to the plate and mixed well and put for the index PCR amplification: 72°C for 3 min; 95°C for 30 sec; 12 cycles of {95°C for 10 sec; 55°C for 30 sec; 72°C for 30 sec}; 72°C for 5 min, and 4°C indefinite hold.

The indexed PCR products were pooled in a single tube, and 0.8X Ampure XP purification (Beckman Coulter) was carried out as per the manufacturer’s recommendation, and finally, the sequencing library was eluted in 35.5 μl of Elution buffer (Qiagen).

SMART seq 2 libraries were sequenced on the NextSeq 500 (Illumina) sequencing platform. Sequencing was done as per the protocol recommendations: paired-end read of 76 bps (read 1), 76 bps (read 2), 8 bps (index 1) and 8 bps (index 2). For a targeted sequencing depth of ∼30 million reads per sample.

Analysis of RNAseq data - Neuron vs EG

Raw reads from neuron- vs. glia-specific FACsorting experiments were processed with the nf-core/rnaseq pipeline (Patel et al., 2020) with default parameters and the BDGP6 reference genome. The STAR-aligned and salmon-quantified expression matrix was then subset to genes known to be enriched in neurons or glia and displayed as a per gene scaled heatmap.

Analysis of RNAseq data - EG-specific differentially expressed genes

After FACsorting of PinkKO-ws vs. control ensheathing glia, library preparation and sequencing resulting FASTQ files were cleaned with fastp (Chen et al., 2018) with default settings. Aligning and counting were carried out with STAR (Dobin et al., 2013) and the 4th 2020 FlyBase Drosophila melanogaster release (r6.35) with the –quantMode flag set as GeneCounts. For differential gene expression testing we used DESeq2 (Love et al., 2014), where we fit a negative binomial model and carried out the Wald-test. We removed the experimental batch as a covariate according to the design formula ∼date+genotype. All genes below a Benjamini-Hochberg corrected p-value of 0.05 were considered deregulated.

ScRNAseq data loading, filtering, clustering, and differential expression analysis

Raw counts data from (Pech et al., 2025), alongside cell annotations provided by the authors, was loaded into Scanpy (v1.9.1) (Wolf et al., 2018) and subset to only include cells in the young control (<= 6-day-old) or Pink1KO-WS samples. After subsetting the data, filtering was performed to remove any cell with less than 250 genes expressed or a percentage of counts coming from mitochondrial genes greater than 15%, the data was then normalized to a total of 10.000 counts and log transformed. Highly variable genes were detected, total_counts and percent of mitochondrial reads were regressed out, and finally the data was scaled to unit variance with a zero mean, clipped to a max of 10.

After data pre-processing, a PCA was performed, Harmony (Korsunsky et al., 2019) was used to correct batches using “sample_id” as a batch key, and 122 components were used for dimensionality reduction and clustering analyses. Leiden clustering was performed with a resolution of 8.0 and using the annotations from (Pech et al., 2025), all cells within each cluster were labeled by taking the most common, original annotation per cluster.

Per cluster, a differential analysis was performed. First, pseudobulk counts were generated by summing the counts per cell from all cells per sample. These counts were then used in DESeq2 (v1.44.0) (Love et al., 2014) comparing Control vs Pink1KO-WS samples. The total number of genes significantly up- or downregulated per cluster (padj <= 0.05 and an abs (log2foldchange) >= 1.5) was counted and plotted.

Statistical analysis

GraphPad Prism was used for visualization and to determine statistical significance. Datasets were tested for normal distribution using the D’Agostino-Person Omnibus and the Shapiro-Wilk normality test. For a normally distributed dataset ordinary one-way ANOVA was used, followed to correct for multiple comparisons by a post hoc Tukey’s test when comparing all the datasets with each other or Dunnett’s test when comparing all the datasets to a general control. For non-normally distributed datasets Kruskal-Wallis test is used, followed by a post hoc Dunn’s test to correct for multiple comparisons. Significance levels are defined as * is p<0.05, ** is p<0.01, *** is p<0.001, and ns, not significant. Effect size is calculated with eta-squared when one-way ANOVA is used, indicated in the figure legend as η. ‘N’ in the legends is used to indicate how many animals were analyzed. Data are plotted as mean ± SD. Specifics on the statistical test used for each analysis are reported in the figure legends.

Key resources table